---
title: "Data integration and preparation: neighbourhood-year panel"
subtitle: "A Predictive Model of Gentrification in Madrid Neighbourhoods"
author: "Eva Lambistos Casado"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "Contents"
    smooth-scroll: true
    number-sections: true
editor: visual
---

## Set-up

```{r set up 1, warning=FALSE, message=FALSE}
options(scipen = 999)
```

```{r set up 2, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(igraph)
library(sf)
```

## Source 1: census tract → neighbourhood bridge table

**Dataset comments**:

This block builds the relationship between census tracts and neighbourhoods. The neighbourhood is a stable unit of analysis for the whole study period except for three known exceptions dated in 2017 (see Note). Census tracts, on the other hand, are not stable over time, so a fixed crosswalk loses a growing fraction of population the further away the crosswalk's reference date is from 2024 (empirically checked against `composición_hogares/` 2017-2025).

Note: 2017 has 128 neighbourhoods, not 131: Ensanche de Vallecas, Valderrivas and El Cañaveral did not exist as neighbourhoods of their own as of January 1st, 2017 - their census tracts belonged to Villa de Vallecas/Vicálvaro.

### Dataset cleaning

#### Source 1a: neighbourhood names (`seccion_censal_barrio_correspondencia.xls`)

`seccion_censal_barrio_correspondencia.xls` is used only to obtain district/neighbourhood names (not to build the tract-neighbourhood relationship, since it has two transcription omissions: tracts 10220 and 16137)

```{r fuente 1a, message=FALSE}
nbsp <- " " # Stores in a variable the NBSP character (non-breaking space) since many cells carry it by default.

raw_corr <-
  read_excel("../data/raw/seccion_censal_barrio_correspondencia.xls", sheet = 1,
             col_names = FALSE) %>% # marks that the first column is data, not a header
  rename(etiqueta = ...1) %>% # renames that first column ...1 to "etiqueta".
  mutate(
      etiqueta = str_squish(str_replace_all(etiqueta, nbsp, " ")),
      # str_replace_all replaces every NBSP with a normal space and str_squish trims spaces and collapses multiple internal spaces into one.

      # main difference: district row: "01. CENTRO" (period + space); neighbourhood row: "01.1 Palacio" (period + digit)
      es_distrito = str_detect(etiqueta, "^\\d{2}\\.\\s"),
      es_barrio   = str_detect(etiqueta, "^\\d{2}\\.\\d+\\s"))

distritos <- raw_corr %>%
  filter(es_distrito) %>% 
  transmute( 
    distrito_num = as.integer(str_extract(etiqueta, "^\\d{2}")), # extracts the first two digits and converts them to integer (drops the leading zero, e.g. "01" → 1).
    distrito_nombre = str_remove(etiqueta, "^\\d{2}\\.\\s*") %>% # Removes the "01. " prefix (digits, period, spaces).
      str_replace_all("\\s*\\.\\s*", "-") %>% # turns " . " into a hyphen
      str_to_title() %>% # converts to title case (initial capital on each word): "San Blas-Canillejas".
      str_replace_all("\\bDe\\b", "de") # Since str_to_title() also capitalizes prepositions like "de", this lowercases them again, giving "de la Concepción".
  )

barrios_madrid <- raw_corr %>%
  filter(es_barrio) %>%
  transmute(
    distrito_num = as.integer(str_extract(etiqueta, "^\\d{2}")), 
    barrio_num   = as.integer(str_extract(etiqueta, "(?<=^\\d{2}\\.)\\d+")), # uses a lookbehind (?<=^\\d{2}\\.)\\d+ — extracts the digits that come right after "two digits + period" at the start, without including those two digits or the period in the result.
    barrio_nombre = str_trim(str_remove(etiqueta, "^\\d{2}\\.\\d+\\s*")), # removes the "01.1 " prefix and trims leftover spaces
    barrio_oficial = distrito_num*10 + barrio_num, # building the neighbourhood-district code as it appears in some files
    codigo_barrio = sprintf("%02d%d", distrito_num, barrio_num) # building the neighbourhood-district code as it appears in other  files = DDB format (2-digit district + neighbourhood without zeros), e.g. "014" = Justicia
  ) %>%
  left_join(distritos, by = "distrito_num") %>% # for each row of the neighbourhood table, looks up in distritos the row with the same distrito_num.
  select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio, barrio_oficial) %>%
  arrange(distrito_num, barrio_num)
```

#### Source 1b: sectioning history (`historia_secciones.xls`)

`historia_secciones.xls` is the 1988-2024 history of each tract code: when it was registered or deregistered, which tract it came from if it was born from a split (`procedencia`), which tract it was absorbed into if it was merged (`destino`), and whether it changed neighbourhood while keeping its number (`barrio_antiguo` + `fecha_modificacion`). It is the source of the tract-neighbourhood relationship and allows reconstructing it for any year.

```{r fuente 1b, message=FALSE}
historia <- read_excel("../data/raw/historia_secciones.xls", sheet = 1, col_names = FALSE) %>%
  rename(distrito = ...1, barrio_registrado = ...2, seccion = ...3,
         fecha_alta = ...4, procedencia = ...5, fecha_baja = ...6,
         destino = ...7, fecha_modificacion = ...8,
         barrio_antiguo = ...9, observaciones = ...10) %>%

  mutate(across(everything(), ~ str_squish(str_replace_all(as.character(.x), nbsp, " ")))) %>% # applies the same cleaning function (~ is shorthand for function (.x)) to all columns at once: converts each column to text > replaces NBSP with normal spaces > trims spaces and collapses multiple internal spaces.

  filter(str_detect(distrito, "^\\d+$")) %>% # Keeps only the rows where the distrito column is exclusively digits, so it drops header and footer.

  mutate(
    distrito = as.integer(distrito),
    barrio_registrado = as.integer(barrio_registrado),
    seccion = as.integer(seccion),
    barrio_antiguo = as.integer(barrio_antiguo),
    fecha_alta = as.integer(fecha_alta), # dates are converted to integer because in the excel they are encoded in YYYYMMDD format
    fecha_baja = as.integer(fecha_baja),
    fecha_modificacion = as.integer(fecha_modificacion)
  )
```

#### Neighbourhood-tract at a reference date

```{r neighbourhood vigente}
# Used to answer "given a census tract code, which neighbourhood did it belong to on a specific date?"
crosswalk_a_fecha <- function(historia, fecha_ref) {
  historia %>%

    filter(fecha_alta <= fecha_ref) %>% # keeps only the tracts that already existed at fecha_ref, whose registration date is earlier than the reference date...
    filter(is.na(fecha_baja) | fecha_baja > fecha_ref) %>% # ...and that have not been deregistered (either no deregistration date, OR a deregistration date later than the reference date).

    mutate(
      barrio_oficial = if_else( # determines which neighbourhood each tract belonged to at each fecha_ref
        !is.na(fecha_modificacion) & !is.na(barrio_antiguo) & fecha_ref < fecha_modificacion,
        # if this tract did have a registered neighbourhood change (fecha_modificacion AND barrio_antiguo are not empty), AND the reference date being queried is earlier than when that change happened...
        barrio_antiguo, # TRUE: the change had not happened yet (date earlier than fecha_ref) = returns the previous neighbourhood (from the barrio_antiguo column).
        barrio_registrado  # FALSE: the change has already happened (date later than fecha_ref) = returns the current neighbourhood (from the barrio_registrado column).
    )) %>%
    transmute(distrito, seccion, barrio_oficial)
}
```

#### Source 1c: year-by-year tract-neighbourhood crosswalk

The tract-neighbourhood relationship is rebuilt as of January 1st of each year in the study period and joined with `barrios_madrid` to bring in the names.

```{r fuente 1c}
anios_estudio_plus <- 2017:2025

crosswalk_seccion_barrio_por_anio <- map_dfr(anios_estudio_plus, function(anio) { # aplies the function to each year and combines all the results 
  crosswalk_a_fecha(historia, as.integer(paste0(anio, "0101"))) %>% mutate(anio = anio) # pastes the year together with the text "0101" to form a date in YYYYMMDD format, stores it as integer, calls the function, and with mutate adds a new anio column with the value of the year being processed in that iteration.
}) %>%

  left_join(barrios_madrid %>% select(barrio_oficial, barrio_num, distrito_nombre, barrio_nombre, codigo_barrio),
            by = "barrio_oficial") %>% # joins the previous table with barrios_madrid to bring in the names (district and neighbourhood) and codigo_barrio, using barrio_oficial as the key.

  mutate(codigo_seccion = sprintf("%02d%03d", distrito, seccion)) %>% # Creates a census tract code as text, combining district (2 digits) and tract (3 digits) - both zero-padded as needed.

  transmute(anio, distrito_num = distrito, distrito_nombre, barrio_num,
            barrio_nombre, codigo_barrio, barrio_oficial, seccion_num = seccion, codigo_seccion) %>% 
  arrange(anio, distrito_num, barrio_num, seccion_num)
```

### Checks (02).

```{r compro 1, message=FALSE}
# checks how many rows the barrios_madrid table has -> 131
stopifnot(nrow(barrios_madrid) == 131)

# checks that for each row of `historia` the district matches what can be recalculated from barrio_registrado -> internal consistency between two columns that should be aligned
stopifnot(all(historia$distrito == historia$barrio_registrado %/% 10))

# checks how many census tracts the official file has -> 2462
secciones_2024 <- read_excel("../data/raw/secciones_2024.xls", sheet = 1, col_names = FALSE) %>%
  rename(distrito = ...1, barrio_oficial = ...2, seccion = ...3) %>%
  mutate(across(everything(), as.character)) %>%
  filter(str_detect(distrito, "^\\d+$")) %>%
  mutate(across(everything(), as.integer))
stopifnot(nrow(secciones_2024) == 2462)

# checks that the district-tract-neighbourhood combinations are identical between both tables
# anti_join returns the rows of A that have no equivalent row in B: diff1 are the rows the reconstruction generates incorrectly relative to the official table, and diff2 are the rows of the official table that the reconstruction fails to generate.
cw_20241015 <- crosswalk_a_fecha(historia, 20241015)
diff1 <- anti_join(cw_20241015, secciones_2024, by = c("distrito", "seccion", "barrio_oficial"))
diff2 <- anti_join(secciones_2024, cw_20241015, by = c("distrito", "seccion", "barrio_oficial"))
stopifnot(nrow(diff1) == 0, nrow(diff2) == 0)

# checks that no crosswalk row was left without a neighbourhood name (join failure).
stopifnot(sum(is.na(crosswalk_seccion_barrio_por_anio$barrio_nombre)) == 0)

# checks there are no duplicates (e.g. a census tract with two neighbourhoods)
stopifnot(sum(duplicated(crosswalk_seccion_barrio_por_anio %>% select(anio, codigo_seccion))) == 0)

# exploratory: extracts the unique neighbourhoods of 2017 and of 2018, and uses anti_join() to find which neighbourhoods are in 2018 but not in 2017.
barrios_2017 <- crosswalk_seccion_barrio_por_anio %>% filter(anio == 2017) %>% distinct(barrio_oficial)
barrios_2018 <- crosswalk_seccion_barrio_por_anio %>% filter(anio == 2018) %>% distinct(barrio_oficial, barrio_nombre)
barrios_2018 %>% anti_join(barrios_2017, by = "barrio_oficial")
```

### Saving

```{r save 1}
dir.create("../data/processed", showWarnings = FALSE) # creates the data processed folder (it increases reproducibility) 

# Saves the three tables built throughout the script as CSV files in the processed data folder
write_csv(barrios_madrid, "../data/processed/barrios_madrid.csv")
write_csv(historia, "../data/processed/historia_secciones_limpio.csv")
write_csv(crosswalk_seccion_barrio_por_anio, "../data/processed/crosswalk_seccion_barrio_por_anio.csv")
```

### View of the final table

```{r view 1}
#| eval: false
crosswalk_seccion_barrio_por_anio <- read_csv("../data/processed/crosswalk_seccion_barrio_por_anio.csv", show_col_types = FALSE)
View(crosswalk_seccion_barrio_por_anio)
```

## Source 2: household composition (`composición_hogares/`)

**Dataset comments**:

The source is at census-tract level, so it uses the crosswalk. The 9 variables to be built detail different household compositions (detailed in columnas_objetivo) and will be the count and percentage of the number of households with such compositions: A single woman aged 16 to 64 + a single man aged 16 to 64, Two adults aged 16 to 64, with no children, Two adults and one child, Two adults and two children, Two adults and three or more children, An adult woman with one or more children, An adult man with one or more children.

### Dataset cleaning

```{r limpieza 2}
# Lists the column names to use to locate them by text match.

columnas_objetivo <- c(
  "Total",
  "Una mujer sola de 16 a 64 años",
  "Un hombre solo de 16 a 64 años",
  "Dos adultos de 16 a 64 años, sin menores",
  "Dos adultos y un menor",
  "Dos adultos y dos menores",
  "Dos adultos y tres o más menores",
  "Una mujer adulta con uno o más menores",
  "Un hombre adulto con uno o más menores"
)

nombres_finales <- c(
  "total_hogares", "hog_mujer_sola_16_64", "hog_hombre_solo_16_64",
  "hog_dos_adultos_sin_menores", "hog_dos_adultos_1_menor", "hog_dos_adultos_2_menores",
  "hog_dos_adultos_3mas_menores", "hog_mujer_adulta_con_menores", "hog_hombre_adulto_con_menores"
)
```

```{r parse 2}
parse_num <- function(x) {
  x <- as.character(x) # converts x (the column vector) to text so transformations can be applied, then stores it as numeric.
  x <- if_else(x == "..", NA_character_, x) # wherever ".." appears writes NA, and keeps the values of x everywhere else
  x <- str_remove_all(x, "\\.") # removes the thousands-separator periods
  as.numeric(x)
}
```

#### Reading per year

```{r lectura 2}
leer_composicion_anio <- function(anio) { # Defines a function that takes a year (anio) and returns that year's household composition data 

  archivo <- list.files("../data/raw/composición_hogares",
                        pattern = paste0("^", anio, "_hogares\\.xlsx?$"), full.names = TRUE) # Looks within the folder for the file whose name exactly matches the pattern: year followed by _hogares.xlsx or _hogares.xls.
             stopifnot(length(archivo) == 1) # warns if more than one file matches that pattern, or none


  hojas <- excel_sheets(archivo) # lists the names of every sheet that excel file has
  hoja_comp <- hojas[str_detect(str_to_lower(hojas), "composici")] # looks for the one with "composici" in lowercase (saves having to match the accent).
  stopifnot(length(hoja_comp) == 1) # warns if more than one sheet matches that pattern, or none


  raw <- read_excel(archivo, sheet = hoja_comp, col_names = FALSE) # reads the composition sheet, without assuming a header row
  cabecera_limpia <- str_to_lower(str_squish(coalesce(as.character(raw[6, ]), as.character(raw[5, ])))) # extracts row 6 (or failing that row 5) in full (where the column names are), converts it to lowercase text and applies str_squish to clean up spaces - the result is a text vector in order.
  objetivo_limpio <- str_to_lower(str_squish(columnas_objetivo))


  idx <- match(objetivo_limpio, cabecera_limpia)
  names(idx) <- columnas_objetivo # assigns the original name to each of the 9 numbers in the idx vector - in order.

 if (any(is.na(idx))) { # checks whether there is at least one TRUE in the vector generated by is.na(idx)
  stop(sprintf("Year %d: there is a column-matching error: %s",
               anio, paste(columnas_objetivo[is.na(idx)], collapse = "; "))) 
 }

  col_seccion <- str_squish(as.character(raw[[2]])) # col2 = tract code
  es_seccion <- replace_na(str_detect(col_seccion, "^\\d{5}$"), FALSE) # str_detect identifies the rows that are a census tract, discarding header, total or empty rows - the result is a TRUE/FALSE/NA vector. replace_na(..., FALSE) replaces any NA that might come out (if some cell was empty) with FALSE - the result is a TRUE/FALSE vector.

  datos <- raw[es_seccion, ] # filters the raw table keeping only the rows where es_seccion is TRUE (by convention) - the ones that are census tracts.
  codigo_seccion <- col_seccion[es_seccion] # applies the same filter to the col_seccion vector


  valores <- tibble(.rows = nrow(datos)) # creates an empty table with the number of rows that "datos" has
  for (i in seq_along(idx)) {
  valores[[nombres_finales[i]]] <- parse_num(datos[[idx[i]]]) # datos[idx[i]] extracts from "datos" the excel column position corresponding to category number i, cleans it with the parse_num defined earlier, and stores it inside valores under the name corresponding to position i of nombres_finales.
}

  bind_cols(tibble(codigo_seccion = codigo_seccion), valores) %>% mutate(anio = anio) # joins the tract-code table and the values table - one table next to the other - adding the year (anio) column.
}
```

```{r resultados 2, message=FALSE}
resultados_hogares <- map(anios_estudio_plus, leer_composicion_anio) # calls the function - the result of map() is a list of 9 elements.
composicion_seccion <- bind_rows(resultados_hogares) # walks through that list of 9 tables and stacks them one below the other, producing the final table.
```

#### Join with the year-by-year crosswalk

```{r union 2}
# A left_join is done, starting from the household composition table which is joined against crosswalk_seccion_barrio_por_anio

composicion_join <- composicion_seccion %>%
  left_join(
    crosswalk_seccion_barrio_por_anio %>%
      select(anio, codigo_seccion, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
    by = c("anio", "codigo_seccion"))
```

#### Aggregation to neighbourhood level

```{r agreg 2}
# from census-tract level to neighbourhood level
composicion_hogares_barrio_anio <- composicion_join %>%
  group_by(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio) %>%
  summarise(across(all_of(nombres_finales), sum), .groups = "drop") %>% # computes the sum of the values for the census tracts within the neighbourhoods
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 2}
# checks there are no duplicate tracts
stopifnot(sum(duplicated(composicion_seccion %>% select(anio, codigo_seccion))) == 0)

# checks that no census tract is missing relative to the already-validated crosswalk count
secciones_leidas <- composicion_seccion %>% distinct(anio, codigo_seccion) %>% count(anio, name = "n_leidas")
secciones_esperadas <- crosswalk_seccion_barrio_por_anio %>% distinct(anio, codigo_seccion) %>% count(anio, name = "n_esperadas")
comparacion_secciones <- left_join(secciones_leidas, secciones_esperadas, by = "anio")
stopifnot(all(comparacion_secciones$n_leidas == comparacion_secciones$n_esperadas))

# checks that no row of composicion_join was left without its corresponding neighbourhood after the join
stopifnot(sum(is.na(composicion_join$barrio_nombre)) == 0)

# checks how many distinct neighbourhoods appear registered in each year (128 in 2017 and 131 in the rest)
barrios_por_anio_hogares <- composicion_hogares_barrio_anio %>% distinct(anio, barrio_nombre) %>% count(anio, name = "n_barrios")
stopifnot(all(barrios_por_anio_hogares$n_barrios[barrios_por_anio_hogares$anio == 2017] == 128))
stopifnot(all(barrios_por_anio_hogares$n_barrios[barrios_por_anio_hogares$anio != 2017] == 131))
```

### Saving

```{r save 2}
write_csv(composicion_hogares_barrio_anio, "../data/processed/composicion_hogares_barrio_anio.csv")
```

### View of the final table

```{r view 2}
#| eval: false
composicion_hogares_barrio_anio <- read_csv("../data/processed/composicion_hogares_barrio_anio.csv", show_col_types = FALSE)
View(composicion_hogares_barrio_anio)
```

## Source 3: rent (`alquiler_barrio.xlsx`)

**Dataset comments**:

The variables extracted are: average rent for multi-family housing, average rent for single-family housing, number of multi-family housing units, and number of single-family housing units. The source is at census-tract level, so each year it is joined with the crosswalk, with some variations. The `alquiler_barrio.xlsx` file is not always published with the tract geography in effect as of January 1st of the year indicated by the sheet name. There are two mismatch cases: the "2017" sheet already uses tract codes that administratively did not exist until after January 1st, 2017, and the "2023" sheet uses 2022 geography. An empirical check concludes that using the 2018 crosswalk to read the "2017" sheet, and the 2022 crosswalk to read the "2023" sheet, brings tracts left without an assigned neighbourhood down to 0%.

```{r cw 3}
anio_crosswalk_alquiler <- c(
  "2017" = 2018, "2018" = 2018, "2019" = 2019, "2020" = 2020, "2021" = 2021,
  "2022" = 2022, "2023" = 2022)
```

### Dataset cleaning

```{r limpieza 3}
# vector to name the columns in the Excel

columnas_alquiler_objetivo <- c(
  grupo_colectiva = "Renta media vivienda colectiva",
  grupo_unifamiliar = "Renta media vivienda unifamiliar",
  n_colectiva = "Viviendas colectivas",
  n_unifamiliar = "Viviendas unifamiliares"
)

# Converts to text, replaces ".." with NA, converts to number
parse_num_alquiler <- function(x) {
  x <- as.character(x)
  x <- if_else(x == "..", NA_character_, x)
  as.numeric(x)
}
```

#### Reading per year

A prior detection of the fixed column positions to use is carried out. Row and column layout verified across every sheet of the file, knowing that the "data range" starts at row 2.

```{r lectura 3, message=FALSE}
# Reads the sheet corresponding to the year
leer_alquiler_anio <- function(anio) {
  archivo <- "../data/raw/alquiler_barrio.xlsx"
  raw_alquiler <- read_excel(archivo, sheet = as.character(anio), col_names = FALSE) # without assuming an automatic header

# Verification of the variables at fixed positions determined within the excel
# Each stopifnot compares the text of a specific cell (raw[row, column], cleaned with str_squish) against the corresponding entry of the columnas_alquiler_objetivo vector
stopifnot(str_squish(as.character(raw_alquiler[3, 8])) == columnas_alquiler_objetivo["grupo_colectiva"])
stopifnot(str_squish(as.character(raw_alquiler[3, 17])) == columnas_alquiler_objetivo["grupo_unifamiliar"])
stopifnot(str_squish(as.character(raw_alquiler[5, 3])) == columnas_alquiler_objetivo["n_colectiva"])
stopifnot(str_squish(as.character(raw_alquiler[5, 4])) == columnas_alquiler_objetivo["n_unifamiliar"])

# Example: if you are processing sheet "2017", as.character(anio) converts it to text "2017", and anio_crosswalk_map_alquiler[["2017"]] looks up that key in the dictionary and returns the associated value (2018).
anio_cw <- anio_crosswalk_alquiler[[as.character(anio)]]

# Renames each column with a useful name
raw_alquiler <- raw_alquiler %>% rename(
    codigo_seccion = ...1,
    n_viviendas_colectiva = ...3,
    n_viviendas_unifamiliar = ...4,
    renta_colectiva_mediana = ...8,
    renta_unifamiliar_mediana = ...17,) %>%

    mutate(codigo_seccion = str_squish(as.character(codigo_seccion))) %>% # cleans up spaces
    filter(str_detect(codigo_seccion, "^\\d{5}$")) # filters by the 5-digit pattern (tracts), dropping unneeded rows

# Builds the final table with the columns year, crosswalk-year, tract code and 4 cleaned numeric variables.
raw_alquiler %>%
    transmute(
      anio = anio,
      anio_crosswalk = anio_cw,
      codigo_seccion,
      n_viviendas_colectiva = parse_num_alquiler(n_viviendas_colectiva),
      n_viviendas_unifamiliar = parse_num_alquiler(n_viviendas_unifamiliar),
      renta_colectiva_mediana = parse_num_alquiler(renta_colectiva_mediana),
      renta_unifamiliar_mediana = parse_num_alquiler(renta_unifamiliar_mediana),
    )
}

# runs the function for every year and stacks the results
anios_estudio_alquiler <- 2017:2023
alquiler_seccion <- map_dfr(anios_estudio_alquiler, leer_alquiler_anio) # walks through the anios_estudio_alquiler vector and for each one calls the leer_alquiler_anio function. Each call returns its own table and they are stacked directly one below the other.
```

#### Join with the year-by-year crosswalk

```{r union 3}
# A left_join is done, starting from the neighbourhood rent table which is joined against crosswalk_seccion_barrio_por_anio

alquiler_join <- alquiler_seccion %>%
  left_join(
    crosswalk_seccion_barrio_por_anio %>%
      select(anio, codigo_seccion, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
    by = c("anio_crosswalk" = "anio", "codigo_seccion")
  )
```

#### (Weighted) aggregation to neighbourhood level

To summarize several tracts into a single neighbourhood figure, the chosen approach was to compute the weighted mean of the tract-level medians, which gives each tract a "weight" proportional to its size (the number of dwellings). Since individual dwelling-level data is not available to extract their median directly, this measure was chosen to approximate it.

```{r agreg 3}
# Defines the function, with x (the values, the rent medians per tract) and w (the weights, the number of dwellings per tract)

weighted_mean_safe <- function(x, w) {
  # checks to make it robust (safe)
  valido <- !is.na(x) & !is.na(w) & w > 0 # the value cannot be missing & the weight cannot be missing & weight > 0
  if (!any(valido)) return(NA_real_) # if no tract of the neighbourhood is usable returns NA // if at least one is valid, applies the following weighted-mean operation to them:
  sum(x[valido] * w[valido]) / sum(w[valido]) # MEAN: multiplies the value by its weight, sums everything and divides by the sum of the weights.
}


# Aggregation by neighbourhood

alquiler_barrio_anio <- alquiler_join %>%
  group_by(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio) %>% # grouped by several columns to avoid losing information.

  summarise(
    # Dwelling counts are summed directly, without weighting, ignoring NAs when summing.
    n_viviendas_colectiva_barrio = sum(n_viviendas_colectiva, na.rm = TRUE),
    n_viviendas_unifamiliar_barrio = sum(n_viviendas_unifamiliar, na.rm = TRUE),

    # The median rent values have the weighted mean applied to them (function above)
    renta_colectiva_barrio = weighted_mean_safe(renta_colectiva_mediana, n_viviendas_colectiva),
    renta_unifamiliar_barrio = weighted_mean_safe(renta_unifamiliar_mediana, n_viviendas_unifamiliar),
    .groups = "drop"
  ) %>%
  arrange(anio, distrito_num, barrio_num) # Sorts the final table by year, district and neighbourhood
```

### Checks (02).

Tract 10157, which corresponds to the neighbourhood of Campamento in the Latina district, has no data in this `alquiler_barrio` file.

```{r campamento}
seccion_leida_2020 <- alquiler_seccion %>% filter(anio == 2020) %>% distinct(codigo_seccion)
seccion_esperada_2020 <- crosswalk_seccion_barrio_por_anio %>% filter(anio == 2020) %>% distinct(codigo_seccion)
anti_join(seccion_esperada_2020, seccion_leida_2020, by = "codigo_seccion")

crosswalk_seccion_barrio_por_anio %>% filter(codigo_seccion == "10157") %>% distinct(anio, distrito_nombre, barrio_nombre)
```

In `renta_colectiva_barrio`, El Cañaveral in 2017 is NA since that neighbourhood was created in November 2017 (see Source 1), so in its first year it barely had a registered rental stock and the Ayuntamiento censors the price for statistical confidentiality.

```{r cañaveral}
# checks the NAs of El Cañaveral
alquiler_barrio_anio %>%
  filter(is.na(renta_colectiva_barrio)) %>%
  select(anio, barrio_nombre, n_viviendas_colectiva_barrio)
```

`renta_unifamiliar_mediana` has a high rate of non-random NAs, since 73% of the NAs (540 of 736) are censorship due to statistical confidentiality. That is, there are single-family dwellings in that neighbourhood-year but none of its tracts have a published price because the sample within the tract is too small. Even so, the single-family dwelling variables are kept in the initial dataset.

```{r MAR}
# checks why the NAs of `renta_unifamiliar_barrio` are not randomly missing
alquiler_barrio_anio %>%
  mutate(
    motivo_na_unifamiliar = case_when(
      !is.na(renta_unifamiliar_barrio) ~ "valid price",
      n_viviendas_unifamiliar_barrio == 0 ~ "no single-family dwelling (count = 0)",
      TRUE ~ "single-family dwelling exists but price censored (statistical confidentiality)"
    )
  ) %>%
  count(motivo_na_unifamiliar)
```

```{r compro 3}
# checks there are no duplicate tracts
stopifnot(sum(duplicated(alquiler_seccion %>% select(anio, codigo_seccion))) == 0)

# checks that no census tract is missing relative to the already-validated crosswalk count
secciones_leidas_alquiler <- alquiler_seccion %>% distinct(anio, codigo_seccion) %>% count(anio, name = "n_leidas")
secciones_esperadas_cw <- crosswalk_seccion_barrio_por_anio %>% distinct(anio, codigo_seccion) %>% count(anio, name = "n_esperadas")

comparacion_secciones_alquiler <- secciones_leidas_alquiler %>%
  mutate(anio_crosswalk = unname(anio_crosswalk_alquiler[as.character(anio)])) %>%
  left_join(secciones_esperadas_cw, by = c("anio_crosswalk" = "anio")) %>%
  mutate(n_esperadas = n_esperadas - 1) # tract 10157 (Latina, Campamento) never appears in alquiler_barrio.xlsx in any year

stopifnot(all(comparacion_secciones_alquiler$n_leidas == comparacion_secciones_alquiler$n_esperadas))

# checks that no row of alquiler_join was left without its corresponding neighbourhood after the join
stopifnot(sum(is.na(alquiler_join$barrio_nombre)) == 0)

# checks how many distinct neighbourhoods appear registered in each year (131 in all)
barrios_por_anio_alquiler <- alquiler_barrio_anio %>% distinct(anio, barrio_nombre) %>% count(anio, name = "n_barrios")
stopifnot(all(barrios_por_anio_alquiler$n_barrios == 131))
```

### Saving

```{r save 3}
write_csv(alquiler_barrio_anio, "../data/processed/alquiler_barrio_anio.csv")
```

### View of the final table

```{r view 3}
#| eval: false
alquiler_barrio_anio <- read_csv("../data/processed/alquiler_barrio_anio.csv", show_col_types = FALSE)
View(alquiler_barrio_anio)
```

## Source 4: declared housing price (`precio_vivienda_declarado_actualizado.xlsx`)

**Dataset comments**:

The extracted variable is the reported housing price. This dataset is at neighbourhood level and in long format. Coverage rate (source limitation): the file's footer states that the Colegio de Registradores captures between 95.3% (2020) and 98.4% (2019) of the actual transactions each year. NA sentinel: some neighbourhood-years have the price as "0,00"; the file's footer explains that "the figure is only shown per neighbourhood when a minimum of 15 cases is available", this is the NA sentinel for statistical confidentiality.

### Dataset cleaning

```{r liimpieza 4, message=FALSE}
precio_vivienda_declarado_raw <- read_excel("../data/raw/precio_vivienda_declarado_actualizado.xlsx",
                                             sheet = 1, col_names = FALSE) %>% # reads the excel sheet without assuming a header row
  rename(anio = ...1, distrito_txt = ...2, barrio_txt = ...3, precio_txt = ...4) %>% # renames the 4 columns

  filter(str_detect(anio, "^\\d{4}$")) %>% # keeps only the rows where anio is exactly 4 digits, discarding any header row, footnotes, etc.
  filter(barrio_txt != distrito_txt & # discards the rows where the text of column 2 and 3 is equal (district total and city) AND...
           str_detect(barrio_txt, "^\\d{3}\\.\\s")) # ...AND requires barrio_txt to start with "3 digits + period + space"


parse_precio_declarado <- function(x) { # defines the cleaning function
  x <- str_remove_all(x, "\\.") # removes the periods (thousands separator)
  x <- str_replace(x, ",", ".") # replaces the decimal comma with a decimal period
  x <- as.numeric(x) # converts to number
  if_else(x == 0, NA_real_, x) # converts to NA since "0,00" = sentinel for <15 transactions, not a real price.
}

precio_vivienda_declarado_barrio_anio <- precio_vivienda_declarado_raw %>%
  transmute( # creates the final columns, dropping the rest
    anio = as.integer(anio),
    codigo_barrio_vivienda  = str_extract(barrio_txt, "^\\d{3}"), # extracts the first 3 digits from the neighbourhood text.
    precio_vivienda_declarado_barrio = parse_precio_declarado(precio_txt) # applies the cleaning function to the price
  ) %>%

  left_join(barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
            by = c("codigo_barrio_vivienda" = "codigo_barrio")) %>% # joins this table with barrios_madrid, bringing in the rest of the neighbourhood information, matching on codigo_barrio (already created).

  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio = codigo_barrio_vivienda, precio_vivienda_declarado_barrio) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

Aeropuerto (Barajas), Atocha, El Cañaveral and El Plantío concentrate the NAs, since they are neighbourhoods with low residential density and/or recently created.

```{r compro 4 NA}
# Where the NAs of `precio_vivienda_declarado_barrio` are concentrated
cat("NA in precio_vivienda_declarado_barrio:",
    sum(is.na(precio_vivienda_declarado_barrio_anio$precio_vivienda_declarado_barrio)), "out of",
    nrow(precio_vivienda_declarado_barrio_anio), "\n")

precio_vivienda_declarado_barrio_anio %>%
  filter(is.na(precio_vivienda_declarado_barrio)) %>%
  count(barrio_nombre, name = "n_anios_con_NA") %>%
  arrange(desc(n_anios_con_NA))
```

```{r compro 4}
# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(precio_vivienda_declarado_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks how many distinct neighbourhoods appear registered in each year (131 in all)
barrios_por_anio_precio <- precio_vivienda_declarado_barrio_anio %>% count(anio, name = "n_barrios")
stopifnot(all(barrios_por_anio_precio$n_barrios == 131))

# checks that no codigo_barrio was left without finding its neighbourhood when joining with barrios_madrid
stopifnot(sum(is.na(precio_vivienda_declarado_barrio_anio$barrio_nombre)) == 0)
```

### Saving

```{r save 4}
write_csv(precio_vivienda_declarado_barrio_anio, "../data/processed/precio_vivienda_declarado_barrio_anio.csv")
```

### View of the final table

```{r view 4}
#| eval: false
precio_vivienda_declarado_barrio_anio <- read_csv("../data/processed/precio_vivienda_declarado_barrio_anio.csv", show_col_types = FALSE)
View(precio_vivienda_declarado_barrio_anio)
```

## Source 5: income (`rentas_barrio.csv`)

**Dataset comments**:

The extracted variables are average gross household income, average gross income per person average net household income and average net income per person. This dataset is at neighbourhood level and in long format. Confirmed directly in the raw file, it was found that at neighbourhood level, `Renta bruta media por hogar` is identical to `Renta neta media por hogar` in every row. At district and city level they are indeed different. `Renta bruta/neta media por persona` are correctly differentiated at neighbourhood level.

### Dataset cleaning

```{r limpieza 5, message=FALSE, warning=FALSE}
rentas_raw <- read_delim("../data/raw/rentas_barrio.csv", delim = ";", 
                          locale = locale(encoding = "UTF-8"),
                          skip = 4, col_names = TRUE, # the header starts after skipping 4 rows
                          col_types = cols(.default = col_character())) |> # forces the columns to text; otherwise the "." (from "21.570") would be read as a decimal and we want it to arrive intact before applying the parse cleaning

rename(anio_txt = ...1, distrito_txt = ...2, barrio_txt = ...3) # renaming the first 3 columns

rentas_datos <- rentas_raw %>%
  filter(str_detect(anio_txt, "^\\d{4}$")) %>% # only rows where anio_txt is 4 digits
  filter(as.integer(anio_txt) <= 2023) %>% # only the ones ≤2023
  filter(str_detect(barrio_txt, "^\\d{3}\\.\\s")) # only rows where barrio_txt has the "3 digits + period + space" pattern at the start, discarding the district-total and city rows.

parse_renta <- function(x) as.numeric(str_remove_all(x, "\\.")) # since these are integers with no decimals, only the thousands-separator periods are removed

rentas_barrio_anio <- rentas_datos %>% # builds the final columns (cleaning the originals with parse_renta)
  transmute(
    anio = as.integer(anio_txt),
    codigo_barrio = str_extract(barrio_txt, "^\\d{3}"), # codigo_barrio extracted from the first 3 digits
    renta_neta_media_persona = parse_renta(`Renta neta media por persona`),
    renta_neta_media_hogar = parse_renta(`Renta neta media por hogar`),
    renta_bruta_media_persona = parse_renta(`Renta bruta media por persona`),
    renta_bruta_media_hogar = parse_renta(`Renta bruta media por hogar`)
  ) %>%

  left_join(barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
            by = "codigo_barrio") %>% # joins this table with barrios_madrid, bringing in the rest of the neighbourhood information, matching on codigo_barrio (already created)
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio, renta_neta_media_persona,
         renta_neta_media_hogar, renta_bruta_media_persona, renta_bruta_media_hogar) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 5}
# checks how many distinct neighbourhoods appear registered in each year (131 in all)
barrios_por_anio_rentas <- rentas_barrio_anio %>% count(anio, name = "n_barrios")
stopifnot(all(barrios_por_anio_rentas$n_barrios == 131))

# checks that no codigo_barrio was left without finding its neighbourhood when joining with barrios_madrid
stopifnot(sum(is.na(rentas_barrio_anio$barrio_nombre)) == 0)

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(rentas_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks whether there are NAs - since this dataset has no declared sentinel.
stopifnot(sum(is.na(select(rentas_barrio_anio, starts_with("renta")))) == 0)

# checks that gross income is never lower than net income (gross >= net)
check_bruta_neta <- function(bruta, neta) {
  stopifnot(sum(bruta < neta, na.rm = TRUE) == 0)
}
check_bruta_neta(rentas_barrio_anio$renta_bruta_media_persona, rentas_barrio_anio$renta_neta_media_persona)
check_bruta_neta(rentas_barrio_anio$renta_bruta_media_hogar, rentas_barrio_anio$renta_neta_media_hogar)
```

### Saving

```{r save 5}
write_csv(rentas_barrio_anio, "../data/processed/rentas_barrio_anio.csv")
```

### View of the final table

```{r view 5}
#| eval: false
rentas_barrio_anio <- read_csv("../data/processed/rentas_barrio_anio.csv", show_col_types = FALSE)
View(rentas_barrio_anio)
```

## Source 6: age bracket (`tramo_edad/`)

**Dataset comments**:

This dataset is at neighbourhood level. From this source only the age indicator of the gentrification index is needed (% population aged 25-39). The results are computed as a value from 0 to 1.


### Dataset cleaning

```{r parse 6}
parse_num_edad <- function(x) as.numeric(str_remove_all(x, "\\.")) # removes the thousands-separator period.
```

#### Reading per year

```{r lectura 6, message=FALSE}
leer_tramo_edad_anio <- function(anio) { # Defines a function that takes a year (anio) and returns that year's age-bracket data already processed.
  archivo <- list.files("../data/raw/tramo_edad",
                         pattern = paste0("^", anio, "_tramo_edad\\.xlsx$"), full.names = TRUE) # Looks within the folder for the file whose name exactly matches the pattern: year followed by _tramo_edad.xlsx or _tramo_edad.xls.
  stopifnot(length(archivo) == 1) # warns if more than one file matches that pattern, or none.


  raw_edad <- read_excel(archivo, sheet = 1, col_names = FALSE) # reads the first sheet without assuming an automatic header.

  raw_edad %>%
    rename(distrito_txt = ...1, barrio_txt = ...2, tramo = ...3, poblacion_col = ...4) %>% # Renames the first 4 columns by position
    filter(distrito_txt != "Total" & !(barrio_txt %in% c("Total", "NO CONSTA"))) %>% # discards the city block, the district-total blocks and "no consta" (population without an assigned neighbourhood)
    filter(tramo %in% c("Total", "25 - 29", "30 - 34", "35 - 39")) %>% 
    transmute( # Builds that year's final table
      anio = anio,
      distrito_txt = str_squish(distrito_txt), # cleans up spaces
      barrio_txt = str_squish(barrio_txt),
      tramo = case_match(tramo, "Total" ~ "poblacion_total_barrio", "25 - 29" ~ "pob_25_29", "30 - 34" ~ "pob_30_34", "35 - 39" ~ "pob_35_39"), 
      poblacion = parse_num_edad(poblacion_col)) # cleans and converts the population to a number.
}

# runs the function for every year and stacks the results
anios_estudio_edad <- 2017:2023
tramo_edad_largo <- map_dfr(anios_estudio_edad, leer_tramo_edad_anio) # walks through the anios_estudio_edad vector and for each one calls the leer_tramo_edad_anio function.
```

#### To wide format and computing the indicator

```{r ancho 6}
tramo_edad_ancho <- tramo_edad_largo %>%
  pivot_wider(names_from = tramo, values_from = poblacion) %>% # one row per neighbourhood-year, one column per bracket
  mutate(
    pob_25_39_barrio = pob_25_29 + pob_30_34 + pob_35_39,
    pct_pob_25_39 = pob_25_39_barrio / poblacion_total_barrio # and its % of the total is computed
  ) %>%
  select(anio, distrito_txt, barrio_txt, poblacion_total_barrio, pob_25_39_barrio, pct_pob_25_39)
```

#### Join by normalized name

This source does not carry a neighbourhood code, only uppercase text with separators different from those of `barrios_madrid` (e.g. `"FUENCARRAL - EL PARDO"` vs `"Fuencarral-El Pardo"`), so the join is done by name, normalizing this table and `barrios_madrid`.

```{r union 6}
# stringi is already installed as a stringr dependency (part of tidyverse)
# defines a name-normalization function

normalizar_nombre <- function(x) {
  x %>%
    stringi::stri_trans_general("Latin-ASCII") %>% # removes accents/diaeresis/ñ -> n
    str_to_upper() %>% # converts everything to uppercase
    str_replace_all("[^A-Z0-9]+", " ") %>% # any character (or run of one or more) that is ^ NOT an uppercase letter (A-Z) nor a digit (0-9) is replaced with a space " ".
    str_squish() # cleans up leftover leading/trailing or duplicate spaces just in case
}

# Applies the normalization function to barrios_madrid, creating two new columns (distrito_key, barrio_key) with the normalized versions of the names
barrios_madrid_norm <- barrios_madrid %>%
  mutate(distrito_key = normalizar_nombre(distrito_nombre),
         barrio_key = normalizar_nombre(barrio_nombre))

# Applies the same normalization to tramo_edad_ancho, and joins both tables using the normalized columns (distrito_key, barrio_key) as the key
tramo_edad_join <- tramo_edad_ancho %>%
  mutate(distrito_key = normalizar_nombre(distrito_txt),
         barrio_key = normalizar_nombre(barrio_txt)) %>%
  left_join(
    barrios_madrid_norm %>%
      select(distrito_key, barrio_key, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
    by = c("distrito_key", "barrio_key")
  )

# Final table: selects the final columns (dropping the original text columns and the normalization keys) and sorts
tramo_edad_barrio_anio <- tramo_edad_join %>%
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         poblacion_total_barrio, pob_25_39_barrio, pct_pob_25_39) %>% 
  arrange(anio, distrito_num, barrio_num)
```

#### Correction of total population for 3 neighbourhoods

The 2018 data is unreliable for Casco Histórico de Vicálvaro, Valdebernardo and Valderrivas, since poblacion_total_barrio shows extreme values that year (69,994, 7 and 6 respectively). Therefore, these values will be replaced with the 2018 total population figures found in the 2019 edition of the indicator panel. Valdebernardo has no data for the "30 - 34" row nor "35 - 39", while Valderrivas has no data for the "30 - 34" row and Casco Histórico has a figure (possibly) calculated over the 69,994 total. For this reason, and since there is no reliable basis to compute the actual 2018 breakdown, these will be set to NA.

```{r}
# identifies the rows generating NAs
tramo_edad_barrio_anio %>%
  filter(is.na(pct_pob_25_39)) %>%
  select(anio, barrio_nombre, poblacion_total_barrio, pob_25_39_barrio, pct_pob_25_39)

# a breakdown problem, not a magnitude problem
tramo_edad_barrio_anio %>%
  filter(distrito_num == 19) %>%
  group_by(anio) %>%
  summarise(poblacion_total_distrito = sum(poblacion_total_barrio), .groups = "drop")

# tribble detailing the actual population values for 2018
poblacion_2018_vicalvaro_real <- tribble(
  ~codigo_barrio, ~poblacion_total_barrio_real,
  "191", 34200, # Casco Histórico de Vicálvaro
  "192", 17637, # Valdebernardo
  "193", 18180  # Valderrivas
) # values from panel_indicadores, 2019 edition (actual year 2018), row "Sexo de la población"

barrios_afectados_2018_vicalvaro <- poblacion_2018_vicalvaro_real$codigo_barrio

tramo_edad_barrio_anio <- tramo_edad_barrio_anio %>%
  left_join(poblacion_2018_vicalvaro_real, by = "codigo_barrio") %>%
  mutate(
    corregir_esta_fila = anio == 2018 & codigo_barrio %in% barrios_afectados_2018_vicalvaro,
    poblacion_total_barrio = if_else(corregir_esta_fila, poblacion_total_barrio_real, poblacion_total_barrio), #  if the row is 2018 and is one of the affected neighbourhoods, the neighbourhood's actual total population is substituted.
    pob_25_39_barrio = if_else(corregir_esta_fila, NA_real_, pob_25_39_barrio), # renamed 
    pct_pob_25_39 = if_else(corregir_esta_fila, NA_real_, pct_pob_25_39) 
  ) %>%
  select(-poblacion_total_barrio_real, -corregir_esta_fila)

# check
tramo_edad_barrio_anio %>%
  filter(anio == 2018, codigo_barrio %in% barrios_afectados_2018_vicalvaro) %>%
  select(codigo_barrio, barrio_nombre, poblacion_total_barrio, pob_25_39_barrio, pct_pob_25_39)
```

### Checks (02).

```{r compro 6 sí}
# checks there are no neighbourhoods left unjoined against barrios_madrid_norm
stopifnot(sum(is.na(tramo_edad_join$codigo_barrio)) == 0)

# checks the number of neighbourhoods per year (128 in 2017, 131 in the rest)
conteo_barrios_edad <- tramo_edad_join %>% count(anio)
stopifnot(conteo_barrios_edad$n[conteo_barrios_edad$anio == 2017] == 128)
stopifnot(all(conteo_barrios_edad$n[conteo_barrios_edad$anio != 2017] == 131))

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(tramo_edad_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks that the non-NA values of pct_pob_25_39 are in the 0-1 range 
stopifnot(all(tramo_edad_barrio_anio$pct_pob_25_39 >= 0 & tramo_edad_barrio_anio$pct_pob_25_39 < 1, na.rm = TRUE))
```

### Saving

```{r save 6}
write_csv(tramo_edad_barrio_anio, "../data/processed/tramo_edad_barrio_anio.csv")
```

### View of the final table

```{r view 6}
#| eval: false
tramo_edad_barrio_anio <- read_csv("../data/processed/tramo_edad_barrio_anio.csv", show_col_types = FALSE)
View(tramo_edad_barrio_anio)
```

## Source 7: housing transactions (`transacciones_vivienda.csv`)

**Dataset comments**:

The variable extracted is rate of housing transactions per inehabitant in each neighbourhood. This dataset is at neighbourhood level and in long format. Coverage rate (source limitation): the file's footer states that the Colegio de Registradores captures between 95.3% (2020) and 98.4% (2019) of the actual transactions each year.

### Dataset cleaning

```{r limpieza 7, warning=FALSE, message=FALSE}
transacciones_raw <- read_delim("../data/raw/transacciones_vivienda.csv", delim = ";", 
                          locale = locale(encoding = "UTF-8"),
                          skip = 5, col_names = TRUE, show_col_types = FALSE, # header after skipping 5 rows
                          col_types = cols(.default = col_character())) %>% # forces text: otherwise "." is read as decimal
  rename(anio_txt = ...1, distrito_txt = ...2, barrio_txt = ...3, transacciones_txt = Total) # renaming 4 columns


transacciones_datos <- transacciones_raw %>%
  filter(str_detect(anio_txt, "^\\d{4}$")) %>% # keeps only the rows where anio_txt is 4 digits
  filter(as.integer(anio_txt) <= 2023) %>% # discards years after 2023
  filter(barrio_txt != distrito_txt) %>% # discards the rows where the neighbourhood text is identical to the district text -> discards "Ciudad de Madrid" and district-total.
  filter(str_detect(barrio_txt, "^\\d{3}\\.\\s")) # only real neighbourhoods with the "3 digits + period + space" pattern (discards "00. No consta")


parse_transacciones <- function(x) {
  x <- if_else(x == "..", NA_character_, x) # replaces the statistical-confidentiality sentinel with NA
  as.numeric(str_remove_all(x, "\\.")) # removes the thousands-separator trap
}


# Creates the final table with year, neighbourhood code, housing transactions and its rate
transacciones_vivienda_barrio_anio <- transacciones_datos %>%
  transmute(
    anio = as.integer(anio_txt),
    codigo_barrio = str_extract(barrio_txt, "^\\d{3}"), # codigo_barrio extracted from the first 3 digits
    transacciones_vivienda_barrio = parse_transacciones(transacciones_txt) # builds the final columns (cleaning the originals with parse_transacciones)
  ) %>%
  left_join(barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio),
            by = "codigo_barrio") %>% # joins this table with barrios_madrid, bringing in the rest of the neighbourhood information, matching on codigo_barrio (already created)


  # the transaction rate is added so it is comparable across neighbourhoods of different sizes
  left_join(tramo_edad_barrio_anio %>% select(anio, codigo_barrio, poblacion_total_barrio),
            by = c("anio", "codigo_barrio")) %>%
  mutate(tasa_transacciones_vivienda = transacciones_vivienda_barrio / poblacion_total_barrio) %>% # computes the rate
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         transacciones_vivienda_barrio, tasa_transacciones_vivienda) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 7}
# checks there is no neighbourhood left unjoined
stopifnot(sum(is.na(transacciones_vivienda_barrio_anio$codigo_barrio)) == 0)

# checks the number of neighbourhoods per year
stopifnot(all(count(transacciones_vivienda_barrio_anio, anio)$n == 131))

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(transacciones_vivienda_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks that the rate is not negative
stopifnot(all(transacciones_vivienda_barrio_anio$tasa_transacciones_vivienda >= 0, na.rm = TRUE))

# checks that the NAs of the rate are exactly the 3 neighbourhoods of 2017 with no population in Source 6.
barrios_sin_poblacion_transacciones <- transacciones_vivienda_barrio_anio %>% filter(is.na(tasa_transacciones_vivienda))
stopifnot(nrow(barrios_sin_poblacion_transacciones) == 3)
stopifnot(all(barrios_sin_poblacion_transacciones$anio == 2017))
stopifnot(all(sort(barrios_sin_poblacion_transacciones$codigo_barrio) == c("183", "193", "194")))
```

### Saving

```{r save 7}
write_csv(transacciones_vivienda_barrio_anio, "../data/processed/transacciones_vivienda_barrio_anio.csv")
```

### View of the final table

```{r view 7}
#| eval: false
transacciones_vivienda_barrio_anio <- read_csv("../data/processed/transacciones_vivienda_barrio_anio.csv", show_col_types = FALSE)
View(transacciones_vivienda_barrio_anio)
```

## Source 8: indicator panel (`panel_indicadores/`)

**Dataset comments**:

From this source 13 indicators are extracted (2017-2023, neighbourhood level): Density, Total population, Average household size, Single-parent households (woman/man, separately), Total number of households, Aging ratio, Over-aging ratio, Proportion of migrant people, People with Spanish/foreign nationality (count and %) and Higher-education level (count and %).

### Dataset cleaning

#### Reading per year and locating the data range

```{r lectura 81}
# To be able to read any edition of the indicator panel
# Open editions from 2017 to 2024 and iterate over 21 sheets

distritos_hojas_panel <- c("CENTRO", "ARGANZUELA", "RETIRO", "SALAMANCA", "CHAMARTÍN", "TETUÁN", "CHAMBERÍ",
  "FUENCARRAL-EL PARDO", "MONCLOA-ARAVACA", "LATINA", "CARABANCHEL", "USERA", "PUENTE DE VALLECAS",
  "MORATALAZ", "CIUDAD LINEAL", "HORTALEZA", "VILLAVERDE", "VILLA DE VALLECAS", "VICÁLVARO",
  "SAN BLAS-CANILLEJAS", "BARAJAS") # the 21 district sheet names, identical across every edition

archivo_edicion_panel <- function(anio_edicion) {
  archivo <- list.files("../data/raw/panel_indicadores",
                         pattern = paste0("^", anio_edicion, "_panel_indicadores\\.xlsx?$"),
                         full.names = TRUE) # .xls in 2017-2020, .xlsx from 2021 onward. # Looks within the folder for the file whose name exactly matches the pattern.
  stopifnot(length(archivo) == 1) # warns if more than one file matches that pattern, or none
  archivo
}

# Locating the data range with localizar_columnas_panel
# looks at the first rows and columns of the sheet searching for the word "Superficie" (which tells it which column holds the indicator labels) and the word "CIUDAD" (which tells it which column the data block starts in and which row the header with neighbourhood names is on) -> this can't be done with fixed positions since it varies by edition

localizar_columnas_panel <- function(raw) {
  esquina <- as.matrix(raw[1:min(10, nrow(raw)), 1:min(6, ncol(raw))]) # esquina is a small matrix (at most 10 rows × 6 columns) with the raw content of that area of the sheet.
  texto <- apply(esquina, c(1, 2), function(v) if (is.na(v)) NA_character_ else str_squish(as.character(v))) # each cell (1,2) of esquina stays NA text if it is NA, and otherwise is converted to text/has its spaces cleaned.

  pos_ciudad <- which(texto == "CIUDAD", arr.ind = TRUE) # compares each cell of the texto matrix against "CIUDAD" and returns a logical matrix of the same size. With arr.ind = TRUE it returns those positions as (row, column) pairs.
  pos_label  <- which(matrix(str_starts(texto, "Superficie \\("), nrow = nrow(texto)), arr.ind = TRUE) # rebuilds the matrix shape to extract positions after searching for a string with str_starts(), which flattens the results into a vector.

  list( # Builds the output list with the three keys
    label_col   = if (nrow(pos_label)  > 0) pos_label[1, "col"]  else NA_integer_, # takes the column number of the first cell that started with "Superficie" else: if no match was found returns NA
    ciudad_col  = if (nrow(pos_ciudad) > 0) pos_ciudad[1, "col"] else NA_integer_, # takes the column number of the first cell that is "CIUDAD"
    fila_header = if (nrow(pos_ciudad) > 0) pos_ciudad[1, "row"] else NA_integer_ # takes the row number of the first cell that is "CIUDAD"
  )
}
```

#### Numeric parsing, column-pair extraction and row location

Both functions are defined to solve problem a and problem b:

a.  Within the same edition some district sheets store numbers as text in European format and others as native Excel numbers. Also, a single period is ambiguous on its own (e.g.: "1.114"), so it is resolved by knowing in advance whether the indicator is a count (thousands separator) or a rate/proportion (decimal), the type defined in specs_panel.

b.  Each geography (City/District/Neighbourhood) occupies a pair of contiguous columns, but what each of those two columns represents depends on the indicator. Depending on the mode (which reflects the indicator's statistical meaning, decided in advance, in specs_panel) it is decided how to extract it. Position rule common to all three modes: if only one column has content, that is the value that will be used, whether A or B, using the determined mode.

-   mode = "simple", type = "proporcion": the indicator can only be a rate. The value may be in A or in B depending on the edition, and whichever has content is used (priority B, then A). -\> Density, Average age, Household size, aging, over-aging, migrants, unemployment.

-   mode = "simple", type = "conteo": the indicator is indeed an actual count. It is normally in the 2nd column (B); the 1st (A), if it exists, is the % over the overall geography (not comparable across neighbourhoods) and is ignored without being read - unless B is empty, in which case it becomes the value to use. -\> Total population, Number of households, single-parent households.

-   mode = "dual": both columns are useful and are read independently. 1st (A) = % over the geography itself and 2nd (B) = absolute count. If one is missing, only that one is left NA, without affecting the other. -\> Spanish/foreign nationality, education level.

Risk accepted in this extraction logic: The pipeline decides that if both have content, the 2nd is the useful value and the 1st is discarded. This rule is based on an empirical check, both manual and through iterations with Claude Code, however there is a risk that in some specific case the values are reversed.

```{r ayb 7}
# solution to problem a.
# type="proporcion": a period (without a comma) is a decimal and is not removed.
# type="conteo": a period followed by groups of exactly 3 digits IS a thousands separator and IS removed.

parse_num_panel <- function(x, tipo = c("conteo", "proporcion")) { # the types can be conteo or proporcion
  tipo <- match.arg(tipo) # match.arg() checks that tipo is one of those two

  if (is.numeric(x)) return(as.numeric(x)) # if the cell already came as a native Excel number, it stays as is
  x <- as.character(x) # if it wasn't numeric it is converted to text
  if (is.na(x) || x == "") return(NA_real_) # and if it is empty or NA it becomes NA_real_


  # CASE 1: if the text contains a comma it is assumed to be European notation and then...
  if (str_detect(x, ",")) {
    x <- str_remove_all(x, "\\.") # all periods are removed from the text (they are thousands separators)
    x <- str_replace(x, ",", ".") # replaces the comma with a period
    return(as.numeric(x)) # and converts to number
  }

  # CASE 2: if there's no comma but there is a period, it depends on the indicator's type
  if (tipo == "conteo" && str_detect(x, "^\\d{1,3}(\\.\\d{3})+$")) { # when the indicator's type is conteo AND the pattern matches groups of three digits separated by a period.
    return(as.numeric(str_remove_all(x, "\\."))) # the periods are removed from the text (they are thousands separators)
  }

  # CASE 3: if the type is "proporcion" + if it is text in normal format (123) + a count that doesn't match the three-digit pattern.
  as.numeric(x)
}


# solution to problem b.
valor_par_panel <- function(raw, fila, col_inicio, modo = c("simple", "dual"), tipo = NULL) {
  # col.inicio is the column where the A-B pair starts
  # fila is where the indicator is located (already found with localizar_fila_panel)
  modo <- match.arg(modo)

  # reads the two values of the A-B pair and with a_ok and b_ok recreates a boolean flag, indicating whether cell a and cell b have real content.
  a <- raw[[fila, col_inicio]]
  b <- raw[[fila, col_inicio + 1]]
  a_ok <- !is.na(a) && a != ""
  b_ok <- !is.na(b) && b != ""

  # CASE 1, simple: if it has content B is used, if that one is empty A will be used. The earlier parse is used.
  if (modo == "simple") {
    if (b_ok) return(list(valor = parse_num_panel(b, tipo)))
    if (a_ok) return(list(valor = parse_num_panel(a, tipo)))
    return(list(valor = NA_real_))
  }

  # CASE 2, dual: A and B are read completely independently
  list(
    proporcion = if (a_ok) parse_num_panel(a, "proporcion") else NA_real_,
    valor = if (b_ok) parse_num_panel(b, "conteo") else NA_real_
  )
}


# Locating the panel's rows, the indicators

localizar_fila_panel <- function(raw, label_col, patron, excluir = NULL) {
  etiquetas <- raw[[label_col]] # extracts the whole indicator-label column from raw
  etiquetas_sq <- ifelse(!is.na(etiquetas), str_squish(as.character(etiquetas)), NA_character_) # if there is a value it converts it to text and applies space cleanup, otherwise NA.

  match_incluir <- str_detect(etiquetas_sq, regex(patron, ignore_case = TRUE)) # returns a TRUE/FALSE vector indicating which labels to include if they match the pattern (anywhere in their text, case-insensitively).
  match_incluir[is.na(match_incluir)] <- FALSE # sets FALSE in the previous vector wherever it is NA.

  if (!is.null(excluir)) { # in case there is an exclusion pattern (same logic)
    match_excluir <- str_detect(etiquetas_sq, regex(excluir, ignore_case = TRUE))
    match_excluir[is.na(match_excluir)] <- FALSE

    match_incluir <- match_incluir & !match_excluir # labels matching the sought pattern AND not matching the discard pattern are included.
  }
  which(match_incluir) # returns the positional indices where the vector is TRUE.
}
```

#### Neighbourhood-name aliases specific to this source

In this source, name normalization (as done in Source 6) is not enough to get a complete set of matches. These mismatch cases exist:

-   *Article added by the panel*: "Las Acacias"/"La Chopera"/"Las Delicias"/"La Estrella"/"Del Pilar"/"Las Águilas"/"Los Ángeles". Resolved by removing a leading article ("El/La/Los/Las/Del") as a second pass (it cannot be applied to all neighbourhoods in a first pass because it would first need to check whether there is already a match).

-   *Abbreviation*: "Villaverde Alto, C.H. Villaverde" vs "Villaverde Alto, Casco Histórico de Villaverde".

-   *Historical pre-2017-reorganization name* (2017 edition only): "San Andrés" (former name of "Villaverde Alto, Casco Histórico de Villaverde") and "Ambroz" (former name of "Valdebernardo", in Vicálvaro, before the mid-2017 restructuring of Vicálvaro's neighbourhoods) and "Palos de Moguer" (former name of "Palos de la Frontera", in Arganzuela), which also appears in editions after 2017, so it is treated as an alias in every edition, not just 2017.

```{r alias}
# tribble defining the neighbourhood aliases
alias_barrio_panel <- tribble(
  ~distrito_key, ~barrio_key_origen, ~barrio_key_canonico,
  "VICALVARO", "AMBROZ", "VALDEBERNARDO",
  "VILLAVERDE", "SAN ANDRES", "VILLAVERDE ALTO CASCO HISTORICO DE VILLAVERDE",
  "ARGANZUELA", "PALOS DE MOGUER", "PALOS DE LA FRONTERA"
)

# normalizes the indicator panel's names
normalizar_barrio_panel <- function(distrito_key, barrio_key) {
  b <- str_replace_all(barrio_key, "\\bC H\\b", "CASCO HISTORICO DE") # expands the abbreviation (CH as its own token, not as part of a word).
  b <- str_remove(b, "^(EL|LA|LOS|LAS|DEL) ") # removes the leading article ->

  # checks whether the district+neighbourhood combination matches any of the table's three historical aliases
  alias <- alias_barrio_panel$barrio_key_canonico[
    match(paste(distrito_key, barrio_key), paste(alias_barrio_panel$distrito_key, alias_barrio_panel$barrio_key_origen))]
  coalesce(alias, b) # if alias has a value that one is used; otherwise b is used.
}

# normalizes barrios_madrid_norm's names
# necessary so that removing the article doesn't break the join in districts where the canonical name DOES carry an article.
barrios_madrid_norm_panel <- barrios_madrid_norm %>%
  mutate(barrio_key = normalizar_barrio_panel(distrito_key, barrio_key))
```

Checked against 2017: every panel row finds a match, only the 3 neighbourhoods that did not yet exist (Ensanche de Vallecas, Valderrivas, El Cañaveral) are left unpaired on the canonical side.

#### Real year of each indicator, per edition

No single edition works for every indicator equally, so there are three families with different lag (this has been checked manually and with Claude Code):

```{r año real}
# This block builds three lookups that translate "actual year I want" -> "edition number where the data is found"

anio_edicion_poblacion <- list("2017" = 2017, "2018" = 2019, "2019" = 2020, "2021" = 2021, "2022" = 2022, "2023" = 2023)
# no panel edition carries 2020 population/density -> substituted with tramo_edad/ further below.

anio_edicion_hogares <- list("2017" = 2017, "2018" = 2019, "2019" = 2020, "2020" = 2021, "2021" = 2022, "2022" = 2023, "2023" = 2024)
# household size, single-parent households, number of households, aging, over-aging, migrants, nationalities, education level.

ediciones_necesarias <- sort(unique(unlist(c(anio_edicion_poblacion, anio_edicion_hogares))))

ediciones_necesarias # 2017, 2019, 2020, 2021, 2022, 2023, 2024 -> the 2018 one is never needed since it only provides 2016 data.
```

#### Reading the indicator panel editions

```{r lectura 82}
# processes ONE indicator (one panel row) and returns the column(s) that correspond to it
# locates the row, checks that it is unique, extracts the value for each neighbourhood creating one column ("simple" mode) or two ("dual" mode)

procesar_indicador <- function(raw, cols, barrio_cols, nombre_col, spec, anio_edicion, hoja) {
  fila_ind <- localizar_fila_panel(raw, cols$label_col, spec$patron, spec$excluir) # looks up this indicator's row, using the inclusion pattern and the exclusion pattern.

  # checks that exactly one row was found, otherwise stops execution
  if (length(fila_ind) != 1) {
    stop(sprintf("Edition %d, sheet %s, indicator %s: %d rows found (1 expected)",
                  anio_edicion, hoja, nombre_col, length(fila_ind)))
  }

  # for each neighbourhood column extracts the value with valor_par_panel() using modo and tipo
  valores <- map(barrio_cols, ~ valor_par_panel(raw, fila_ind, .x, modo = spec$modo, tipo = spec$tipo))

  # Depending on the mode, generates the function's result: columnas (the variable names) and valores paired by position.
  if (spec$modo == "dual") {
    list(
      columnas = c(paste0(nombre_col, "_pct"), nombre_col),
      valores = list(map_dbl(valores, "proporcion"), map_dbl(valores, "valor")))
  } else {
    list(columnas = nombre_col, valores = list(map_dbl(valores, "valor")))
  }
}


# function that reads one full panel edition (21 sheets) and extracts the indicators
leer_edicion_multiindicador <- function(anio_edicion, specs) {
  archivo <- archivo_edicion_panel(anio_edicion)
  map_dfr(distritos_hojas_panel, function(hoja) { # applies the function and combines all results into a single data frame by stacking rows

    raw <- read_excel(archivo, sheet = hoja, col_names = FALSE) # reads the sheet without a header
    cols <- localizar_columnas_panel(raw) # applies localizar_columnas_panel (data range)
    barrio_cols_all <- seq(cols$ciudad_col + 4, ncol(raw), by = 2) # Computes the columns where each neighbourhood's pair starts, generating a sequence skipping two at a time, because each geography occupies 2 columns. Skips the first 4 columns from the first CIUDAD column, which correspond to the 4 columns of CIUDAD + Distrito (checked manually).

    # for each value of barrio_cols_all extracts the header text: the neighbourhood name
    barrio_nombres_all <- map_chr(barrio_cols_all, ~ {
      v <- raw[[cols$fila_header, .x]]
      if (is.character(v) && !is.na(v)) str_squish(v) else NA_character_ # if there is text it cleans it, otherwise NA
    })

    # Filters the columns that did have a real neighbourhood name. barrio_cols and barrio_nombres end up as the "real column-real name" pairs.
    ok <- !is.na(barrio_nombres_all) & barrio_nombres_all != ""
    barrio_cols <- barrio_cols_all[ok]
    barrio_nombres <- barrio_nombres_all[ok]



    resultado <- tibble(distrito_txt = hoja, barrio_txt = barrio_nombres)
    # for each indicator in specs, it is processed (locate row + extract values) and
    # its resulting columns are added to `resultado` -- the detail of how each
    # indicator is processed lives in procesar_indicador(), defined above
    for (nombre_col in names(specs)) {
      r <- procesar_indicador(raw, cols, barrio_cols, nombre_col, specs[[nombre_col]], anio_edicion, hoja)
      for (i in seq_along(r$columnas)) resultado[[r$columnas[i]]] <- r$valores[[i]] # takes the columns (new variables) and values and adds them to the results table
    }
    resultado
  })
}
```

Cases with text variants across editions, all confirmed against the actual files: - aging/over-aging: "Proporción de..." 2017-2023, "Porcentaje de..." 2024 - migrants (non-OECD/EU): "Proporción de inmigrantes..." 2017-2020, "...personas migrantes..." 2021-2024. - Spanish/foreign nationality: carry their own date in parentheses from 2021 onward, so the pattern optionally allows it. - Number of households: "Nº total de hogares" 2017-2023, "Total hogares" in 2024. - education level: due to various typographical errors the pattern is anchored on "no universitario", which is stable across all of them.

```{r patrones 8}
# Row patterns per indicator: details the pattern to identify its label text and the modo/tipo specifications.

specs_panel <- list(
  densidad = list(patron = "^Densidad \\(", modo = "simple", tipo = "proporcion"),

  poblacion_total_panel = list(patron = "^Sexo de la población$", modo = "simple", tipo = "conteo"),

  tamano_hogar = list(patron = "^Tamaño medio del hogar", modo = "simple", tipo = "proporcion"),

  monop_mujer = list(patron = "^Hogares monoparentales: una mujer adulta", modo = "simple", tipo = "conteo"),

  monop_hombre = list(patron = "^Hogares monoparentales: un hombre adulto", modo = "simple", tipo = "conteo"),

  n_hogares = list(patron = "^(Nº total de hogares|Total hogares)$", modo = "simple", tipo = "conteo"),

  envejecimiento = list(patron = "^(Proporción|Porcentaje) de envejecimiento", modo = "simple", tipo = "proporcion"),

  sobreenvejecimiento = list(patron = "^(Proporción|Porcentaje) de sobre-?envejecimiento", modo = "simple", tipo = "proporcion"),

  migrantes = list(patron = "^Proporción de (inmigrantes|personas migrantes)", modo = "simple", tipo = "proporcion"),

  nac_espanola = list(patron = "^Personas con nacionalidad española(\\s*\\(.*\\))?$", modo = "dual"),

  nac_extranjera = list(patron = "^Personas con nacionalidad extranjera(\\s*\\(.*\\))?$", modo = "dual"),

  nivel_estudios_superiores = list(patron = "estudios superior.*no universitario|no universitario.*estudios superior", modo = "dual")
)

# Only the "hogares" family (see the actual-year tables further below) needs the 2024 edition
specs_panel_familia_hogares <- specs_panel[c("tamano_hogar", "monop_mujer", "monop_hombre", "n_hogares",
  "envejecimiento", "sobreenvejecimiento", "migrantes", "nac_espanola", "nac_extranjera", "nivel_estudios_superiores")]
```

```{r lectura 83, message=FALSE}
# This function takes one full edition already read with leer_edicion_multiindicador() and turns it into a table with codigo_barrio.

leer_edicion_normalizada <- function(anio_edicion, specs) {
  tabla <- leer_edicion_multiindicador(anio_edicion, specs) # calls the previous function building that edition's full table.

  tabla_join <- tabla %>%
    mutate(distrito_key = normalizar_nombre(distrito_txt), barrio_key = normalizar_nombre(barrio_txt)) %>% # creates distrito_key and barrio_key columns by applying normalizar_nombre (created in source 6)

    mutate(barrio_key = normalizar_barrio_panel(distrito_key, barrio_key)) %>% # changes barrio_key by applying the normalizar_barrio_panel function defined earlier.

    left_join(
      barrios_madrid_norm_panel %>%
        select(distrito_key, barrio_key, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio), by = c("distrito_key", "barrio_key")) # joins the table with barrios_madrid_norm_panel -> with already-normalized neighbourhood names.


  n_sin_barrio <- sum(is.na(tabla_join$codigo_barrio)) # counts how many rows of tabla_join were left without codigo_barrio
  if (n_sin_barrio > 0) { # if there is at least one unmatched row, locate where
    stop(sprintf("Edition %d: %d rows without a neighbourhood after normalizing (%s does not cover all cases)", anio_edicion, n_sin_barrio, "normalizar_barrio_panel()"))
  }
  tabla_join %>% select(-distrito_txt, -barrio_txt, -distrito_key, -barrio_key)
}

# reads each of the necessary editions (keeps the editions separate in a list).
datos_por_edicion_panel <- map(ediciones_necesarias, function(ed) {
  specs_usar <- if (ed == 2024) specs_panel_familia_hogares else specs_panel # only the 2024 edition is used exclusively for the "hogares" family, for the rest the specs_panel indicators are looked up
  leer_edicion_normalizada(ed, specs_usar) # reads the edition and normalizes it to codigo_barrio (function above)
})

# replaces the names of datos_por_edicion_panel with what's on the right-hand side ("2017-2024")
names(datos_por_edicion_panel) <- as.character(ediciones_necesarias)
```

#### Assembly by actual year

```{r ensamblado 8}
# This function joins, for each actual year (2017 to 2023), the population columns and the household columns, taking them from whichever edition corresponds to each family.

claves_barrio_panel <- c("distrito_num", "distrito_nombre", "barrio_num", "barrio_nombre", "codigo_barrio") # full neighbourhood identification

# variables belonging to each family (in terms of edition lag)
cols_poblacion <- c("densidad", "poblacion_total_panel")
cols_hogares <- c("tamano_hogar", "monop_mujer", "monop_hombre", "n_hogares",
  "envejecimiento", "sobreenvejecimiento", "migrantes",
  "nac_espanola_pct", "nac_espanola", "nac_extranjera_pct", "nac_extranjera",
  "nivel_estudios_superiores_pct", "nivel_estudios_superiores")

# Joined on the already-normalized codigo_barrio (and the rest of the keys).
ensamblar_anio_real <- function(anio_real) {
  base <- NULL

  ed_pob <- anio_edicion_poblacion[[as.character(anio_real)]] # looks up in the anio_edicion_poblacion table which edition carries the population data for this actual year (if there is no entry, as in 2020, it returns NULL).

  if (!is.null(ed_pob)) { # if it is not NULL... selects the relevant columns
    tabla <- datos_por_edicion_panel[[as.character(ed_pob)]] %>% select(all_of(claves_barrio_panel), all_of(cols_poblacion))
    base <- tabla
  }

  ed_hog <- anio_edicion_hogares[[as.character(anio_real)]] # same logic for the other family
  if (!is.null(ed_hog)) {
    tabla <- datos_por_edicion_panel[[as.character(ed_hog)]] %>% # looks up in the anio_edicion_hogares table which edition carries the population data for this actual year
    select(all_of(claves_barrio_panel), all_of(cols_hogares))
    base <- if (is.null(base)) tabla else full_join(base, tabla, by = claves_barrio_panel)
  }

  base %>% mutate(anio = anio_real, .before = 1) # adds an actual-year column
}
panel_largo <- map_dfr(2017:2023, ensamblar_anio_real) # combines the data with actual years into a single table
```

#### Correction of density and total population variables in 2020

No panel edition carries 2020 population and density, but both figures could be extracted from `tramo_edad_barrio_anio` (Source 6). Density is recomputed with that population divided by the neighbourhood's Surface area (since it is a stable value, 2020 will be used).

```{r sustitucion 8, message=FALSE}
# reuses leer_edicion_normalizada() but with a specs list for the "Superficie" indicator.
leer_superficie_barrio <- function(anio_edicion) {
  leer_edicion_normalizada(anio_edicion, list(
    superficie_ha = list(patron = "^Superficie \\(", modo = "simple", tipo = "proporcion")
  ))
}
superficie_barrio_panel <- leer_superficie_barrio(2020)


sustituto_2020 <- tramo_edad_barrio_anio %>%
  filter(anio == 2020) %>% # Filters only 2020 from Source 6's table
  select(codigo_barrio, poblacion_total_barrio) %>%
  left_join(superficie_barrio_panel %>% select(codigo_barrio, superficie_ha), by = "codigo_barrio") # joins the surface area to that table, matching on codigo_barrio


# checks that no neighbourhood was left without a surface area after the join
stopifnot(sum(is.na(sustituto_2020$superficie_ha)) == 0) # if this fails, 2023 works.

# the result is left with just codigo_barrio, the renamed population, and the computed density.
sustituto_2020 <- sustituto_2020 %>%
  transmute(
    codigo_barrio,
    poblacion_total_panel_sust = poblacion_total_barrio,
    densidad_sust = poblacion_total_barrio / superficie_ha)

# Actual substitution of the density and population variables in 2020
panel_indicadores_barrio_anio  <- panel_largo %>%
  left_join(sustituto_2020, by = "codigo_barrio") %>% # joins the substitute to the panel (with actual year)
  mutate(
    poblacion_total_panel = if_else(anio == 2020, poblacion_total_panel_sust, poblacion_total_panel), # replaces the variables in the rows where the actual year is 2020, and for the other years keeps the value it already had
    densidad = if_else(anio == 2020, densidad_sust, densidad)
  ) %>% select(-poblacion_total_panel_sust, -densidad_sust) # the already-used substitutes are dropped
```

#### Correction of nac_espanola_pct / nac_extranjera_pct

The percentage column for both Spanish and foreign nationality that the indicator panel source itself provides is not reliable, even producing percentages above 100% in some cases (Atocha 2021 = 104.5%, El Cañaveral 2021 = 162.2%). The error also appears in El Cañaveral and Atocha in 2020 and 2022, and in several more neighbourhoods only in 2022 (Ángeles, Quintana, Ventas, San Pascual). Since it has been verified that `nac_espanola + nac_extranjera` matches `poblacion_total_panel` with a median error margin of 0% (90th percentile value of 0, and an absolute maximum of 0.83%), the calculation of both percentages is validated from the count and `poblacion_total_panel`, instead of using the "%" that the source itself provides.

```{r}
# check for percentages > 1 (above 100% in the raw data) -> El Cañaveral and Atocha 2021
panel_indicadores_barrio_anio %>%
  filter(nac_espanola_pct > 1 | nac_extranjera_pct > 1) %>%
  select(anio, barrio_nombre, codigo_barrio, nac_espanola_pct, nac_extranjera_pct)

# pct_calculado is the one extracted from count/total population, while nac_espanola_pct is the % that appears as data in the indicator panel -> there is a difference in several cases.
panel_indicadores_barrio_anio %>%
  mutate(
    pct_calculado = nac_espanola / poblacion_total_panel,
    diferencia = abs(nac_espanola_pct - pct_calculado)
  ) %>%
  arrange(desc(diferencia)) %>%
  select(anio, barrio_nombre, codigo_barrio, nac_espanola_pct, pct_calculado, diferencia,
         nac_espanola, nac_extranjera, poblacion_total_panel) %>%
  slice_head(n = 10)

# check nac_espanola + nac_extranjera = poblacion_total_panel
panel_indicadores_barrio_anio %>%
  mutate(diferencia_relativa = (nac_espanola + nac_extranjera - poblacion_total_panel) / poblacion_total_panel) %>%
  summarise(
    n = n(),
    minimo = min(diferencia_relativa),
    mediana = median(diferencia_relativa),
    p90 = quantile(diferencia_relativa, 0.90), # at least 90% of the 914 rows have zero difference
    maximo = max(diferencia_relativa),
    max_absoluto = max(abs(diferencia_relativa)))
```

```{r correccion nacionalidad}
# substitution of the source's "%" with count/population
panel_indicadores_barrio_anio <- panel_indicadores_barrio_anio %>%
  mutate( #  overwrites the two columns nac_espanola_pct and nac_extranjera_pct
    nac_espanola_pct = nac_espanola / poblacion_total_panel,
    nac_extranjera_pct = nac_extranjera / poblacion_total_panel)
```

#### Correction of the migrants variable in 2023

The `migrantes` variable (non-EU/rest-of-OECD foreigners) is by definition a subset of `nac_extranjera_pct` (all foreigners) and both are computed under the same population total. So, from the subtraction `nac_extranjera_pct - migrantes` one obtains those migrants originating from EU/OECD countries. In 2 neighbourhoods of Arganzuela in 2023 (Acacias, Legazpi) the subtraction is negative. This has been checked against the raw Excel of `2024_panel_indicadores.xlsx` (the edition that uses actual year 2023 for this indicator family), and it is observed that `migrantes` shows a jump in 2023 for both neighbourhoods compared to their previous years, while `nac_extranjera_pct` follows a smooth trend with no jump. The correction is limited to these 2 cases since the rest show data consistent with a genuine rise in the migrantes variable.

```{r correccion migrantes}
# identifies the neighbourhood-years where nac_extranjera_pct - migrantes comes out negative
filas_migrantes_no_fiable <- panel_indicadores_barrio_anio %>%
  mutate(diferencia_ue_ocde = nac_extranjera_pct - migrantes) %>% # subtraction
  filter(diferencia_ue_ocde < 0) %>%
  select(anio, codigo_barrio, barrio_nombre, nac_extranjera_pct, migrantes, diferencia_ue_ocde)
filas_migrantes_no_fiable

# checks that the "migrantes" problem is in Acacias and Legazpi 2023
stopifnot(nrow(filas_migrantes_no_fiable) == 2)
stopifnot(all(filas_migrantes_no_fiable$anio == 2023))
stopifnot(all(sort(filas_migrantes_no_fiable$codigo_barrio) == c("022", "024")))

# migrantes is set to NA only in the rows identified above
panel_indicadores_barrio_anio <- panel_indicadores_barrio_anio %>%
  left_join(filas_migrantes_no_fiable %>% transmute(anio, codigo_barrio, no_fiable = TRUE),
            by = c("anio", "codigo_barrio")) %>%
  mutate(migrantes = if_else(!is.na(no_fiable), NA_real_, migrantes)) %>% # if it falls under unreliable it is set to NA, otherwise the same value is kept.
  select(-no_fiable)

# the subtraction is done -> creating the pct_extranjeros_ue_ocde variable
panel_indicadores_barrio_anio <- panel_indicadores_barrio_anio %>%
  mutate(pct_extranjeros_ue_ocde = nac_extranjera_pct - migrantes)
```

```{r tabla 8}
# Final table
panel_indicadores_barrio_anio <- panel_indicadores_barrio_anio %>%
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         densidad, poblacion_total_panel,
         tamano_hogar, monop_mujer, monop_hombre, n_hogares,
         envejecimiento, sobreenvejecimiento, migrantes, pct_extranjeros_ue_ocde,
         nac_espanola_pct, nac_espanola, nac_extranjera_pct, nac_extranjera,
         nivel_estudios_superiores_pct, nivel_estudios_superiores) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 8}
# checks there is no neighbourhood left unjoined
stopifnot(sum(is.na(panel_largo$codigo_barrio)) == 0)

# checks the number of neighbourhoods per year
barrios_por_anio_panel <- panel_indicadores_barrio_anio %>% count(anio, name = "n_barrios")
stopifnot(barrios_por_anio_panel$n_barrios[barrios_por_anio_panel$anio == 2017] == 128)
stopifnot(all(barrios_por_anio_panel$n_barrios[barrios_por_anio_panel$anio != 2017] == 131))

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(panel_indicadores_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks that proportions/rates are within range (0-1)
cols_proporcion <- c("envejecimiento", "sobreenvejecimiento", "migrantes", "pct_extranjeros_ue_ocde", "nac_espanola_pct", "nac_extranjera_pct", "nivel_estudios_superiores_pct")
fuera_de_rango <- panel_indicadores_barrio_anio %>%
  summarise(across(all_of(cols_proporcion), ~ any(. < 0 | . > 1, na.rm = TRUE)))
stopifnot(all(!fuera_de_rango))

# checks that the only NAs of migrantes/pct_extranjeros_ue_ocde are Acacias and Legazpi 2023
stopifnot(sum(is.na(panel_indicadores_barrio_anio$migrantes)) == 2)
stopifnot(sum(is.na(panel_indicadores_barrio_anio$pct_extranjeros_ue_ocde)) == 2)
```

### Saving

```{r save 8}
write_csv(panel_indicadores_barrio_anio, "../data/processed/panel_indicadores_barrio_anio.csv")
```

### View of the final table

```{r view 8}
#| eval: false
panel_indicadores_barrio_anio <- read_csv("../data/processed/panel_indicadores_barrio_anio.csv", show_col_types = FALSE)
View(panel_indicadores_barrio_anio)
```

## Source 9: census of business premises (`censo_locales/`)

**Dataset comments**:

The following CNAE Divisions were chosen to build the variables, which deal only with "Abierto" (open) premises. There are two manual exclusions within 55 and 56, decided so those two groups are purer proxies of "tourism" and of "bars and restaurants". `559001` (Colegios Mayores and student halls of residence) is excluded from 55. `562901`-`562905` (canteens in sports facilities/offices, education centers, senior centers, social service centers, hospitals) are excluded from 56 since they are not restaurant establishments open to the public.

| Group | CNAE Divisions | Given name |
|------------------------|------------------------|------------------------|
| `comercio_menor` | 47 | Retail trade |
| `turismo_alojamiento` | 55 (minus `559001`) | Tourism and accommodation |
| `bares_restaurantes` | 56 (minus `562901`-`562905`) | Bars and restaurants |
| `inmobiliarias` | 68 | Real estate activities |
| `educacion` | 85 | Education |
| `sanitarias` | 86 | Health |
| `servicios_sociales` | 87 + 88 | Social services |
| `arte_recreo_ocio` | 90 + 91 + 93 | Arts, recreation and entertainment |

The variables extracted are the count and percentage of these establishment activities.

### Dataset cleaning

```{r limpieza 9, warning=FALSE}
encoding_censo_por_anio <- c( # it has been confirmed that each file has a different encoding
  "2017" = "WINDOWS-1252", "2018" = "UTF-8", "2019" = "WINDOWS-1252",
  "2020" = "WINDOWS-1252", "2021" = "WINDOWS-1252", "2022" = "UTF-8", "2023" = "UTF-8"
)

# `id_barrio_local` is not `codigo_barrio`, so codigo_barrio is reconstructed (format "DDB", e.g. "083") from id_barrio_local (format district without a leading zero + 2-digit neighbourhood, e.g. "803")
id_barrio_local_a_codigo_barrio <- function(id_barrio_local) {
  id_num <- as.integer(id_barrio_local)
  distrito_num <- id_num %/% 100 # returns the district
  barrio_num <- id_num %% 100 # returns the neighbourhood
  sprintf("%02d%d", distrito_num, barrio_num) # rebuilds the codigo_barrio format
}

# function that reads the census of business premises
leer_censo_locales_raw_anio <- function(anio) {
  archivo <- sprintf("../data/raw/censo_locales/%d_locales.csv", anio)
  encoding <- encoding_censo_por_anio[[as.character(anio)]] # selects the encoding

# reads the file applying the encoding and forcing every column to be read as text (to prevent id_epigrafe from being read as a number and losing digits).
  raw <- read_delim(archivo, delim = ";",
                     locale = locale(encoding = encoding),
                     col_types = cols(.default = col_character()),
                     show_col_types = FALSE)

  raw %>%
    transmute(
      anio = anio,
      codigo_barrio = id_barrio_local_a_codigo_barrio(id_barrio_local), # applies the reconstruction
      desc_situacion_local,
      id_division = if_else(str_detect(id_division, "^COMERCIO AL POR MENOR"), "47", str_squish(id_division)), # fixes the single data-corruption case that falls within the 8 groups of interest
      id_epigrafe = str_squish(id_epigrafe))
}

# Applied for every year
censo_locales_raw_todos <- map_dfr(2017:2023, leer_censo_locales_raw_anio)

# censo_locales_todos is created by keeping only the open premises
censo_locales_todos <- censo_locales_raw_todos %>%
  filter(desc_situacion_local == "Abierto") %>%
  select(anio, codigo_barrio, id_division, id_epigrafe)

# looks at the neighbourhoods that have premises records in the table
barrios_presentes_todos <- censo_locales_raw_todos %>% distinct(anio, codigo_barrio)
```

#### Economic-activity groups and filtering

```{r econom 9}
# definition of the economic-activity groups by Division level
division_a_grupo_econom <- tribble(
  ~id_division, ~grupo_comercial,
  "47", "comercio_menor",
  "55", "turismo_alojamiento",
  "56", "bares_restaurantes",
  "68", "inmobiliarias",
  "85", "educacion",
  "86", "sanitarias",
  "87", "servicios_sociales",
  "88", "servicios_sociales",
  "90", "arte_recreo_ocio",
  "91", "arte_recreo_ocio",
  "93", "arte_recreo_ocio"
)
cols_grupo <- unique(division_a_grupo_econom$grupo_comercial) # the 8 final groups

# definition of the manual exclusions of divisions 55 and 56
epigrafes_excluidos_manual <- c(
  "559001",
  "562901", "562902", "562903", "562904", "562905")

# denominator: total active premises per neighbourhood-year, over ALL divisions (not just the chosen ones)
total_locales_barrio_anio <- censo_locales_todos %>%
  count(anio, codigo_barrio, name = "n_locales_activos_total")

# numerator: only the premises from the 8 chosen divisions/groups, removing exclusions
conteo_grupo_barrio_anio <- censo_locales_todos %>%
  filter(!(id_epigrafe %in% epigrafes_excluidos_manual)) %>% # excludes epigraphs
  inner_join(division_a_grupo_econom, by = "id_division") %>% # keeps the chosen ones
  count(anio, codigo_barrio, grupo_comercial, name = "n") %>% # counts premises per neighbourhood-year-group
  pivot_wider(names_from = grupo_comercial, values_from = n, names_prefix = "n_", values_fill = 0) # converts that long format to wide, creating one column per economic group.
```

#### Neighbourhood × year grid and assembly

A real neighbourhood with no registered premises that year (Valderrivas/El Cañaveral in 2018) would disappear without generating any row. Because of this, a template of which neighbourhoods exist in which year is built (exception of 2017 with 128 neighbourhoods), and the count is joined to it even if it is 0. In this case, the only neighbourhoods with no row at all in the raw data are Valderrivas and El Cañaveral (`barrios_sin_cobertura`) since they are neighbourhoods created recently in 2018, so it is possible that, whether due to economic activity not yet being located in that neighbourhood or due to it not yet being recorded in an emerging area, there is no data about it.

```{r grid 9}
barrios_sin_entidad_2017 <- c("183", "193", "194") # neighbourhoods that did not yet exist in 2017

# removes those neighbourhoods only when the year is 2017, the rest stays the same.
grid_censo_barrio_anio <- map_dfr(2017:2023, function(anio) {
  base <- barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio)
  if (anio == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017)
  base %>% mutate(anio = anio, .before = 1)
})

# computes the neighbourhoods without coverage
barrios_sin_cobertura <- grid_censo_barrio_anio %>%
  select(anio, codigo_barrio) %>%
  anti_join(barrios_presentes_todos, by = c("anio", "codigo_barrio"))


# Final assembly
censo_locales_barrio_anio <- grid_censo_barrio_anio %>%
  left_join(total_locales_barrio_anio, by = c("anio", "codigo_barrio")) %>%

  left_join(barrios_sin_cobertura %>% mutate(sin_cobertura = TRUE), by = c("anio", "codigo_barrio")) %>%
  mutate(
    sin_cobertura = replace_na(sin_cobertura, FALSE),
    n_locales_activos_total = case_when(
      !is.na(n_locales_activos_total) ~ n_locales_activos_total, # active = at least 1 open premise
      sin_cobertura              ~ 0, 
      TRUE                       ~ 0 # 0 rows in the raw data
    )
  ) %>%
  
  # Same pattern for the 8 per-group columns
  left_join(conteo_grupo_barrio_anio, by = c("anio", "codigo_barrio")) %>%
  mutate(across(all_of(paste0("n_", cols_grupo)),
                ~ case_when(
                    is.na(n_locales_activos_total) ~ NA_real_,
                    n_locales_activos_total == 0    ~ 0,
                    TRUE                             ~ replace_na(.x, 0)
                ))) %>%

  mutate(across(all_of(paste0("n_", cols_grupo)),
                ~ case_when(
                    sin_cobertura                 ~ 0, 
                    n_locales_activos_total == 0  ~ NA_real_, 
                    TRUE                           ~ .x / n_locales_activos_total
                ),
                .names = "pct_{sub('^n_', '', .col)}")) %>% # if sin_cobertura, the result is 0; if n_locales_activos_total is 0 for another reason, the result is NA (undefined, 0/0); otherwise, computes the group's share of total active premises.
  select(-sin_cobertura) %>%
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         all_of(paste0("n_", cols_grupo)), all_of(paste0("pct_", cols_grupo)),
         n_locales_activos_total) %>%
  arrange(anio, distrito_num, barrio_num)
```

#### Premises in residential use

```{r uso vivienda 9}
# % of premises in residential use
total_locales_todos_estados <- censo_locales_raw_todos %>%
  count(anio, codigo_barrio, name = "n_locales_todos_estados") # counts every kind of premise

conteo_uso_vivienda <- censo_locales_raw_todos %>%
  filter(desc_situacion_local == "Uso vivienda") %>%
  count(anio, codigo_barrio, name = "n_uso_vivienda") # counts the premises used as housing

# Same logic as the previous case
uso_vivienda_barrio_anio <- grid_censo_barrio_anio %>%
  left_join(total_locales_todos_estados, by = c("anio", "codigo_barrio")) %>% # adds n_locales_todos_estados
  left_join(conteo_uso_vivienda, by = c("anio", "codigo_barrio")) %>% # adds n_uso_vivienda
  left_join(barrios_sin_cobertura %>% mutate(sin_cobertura = TRUE), by = c("anio", "codigo_barrio")) %>% # flag for cases with or without coverage
  
  mutate(
    sin_cobertura = replace_na(sin_cobertura, FALSE),
    n_locales_todos_estados = case_when(
      !is.na(n_locales_todos_estados) ~ n_locales_todos_estados,
      sin_cobertura ~ 0,
      TRUE ~ 0
    ),
    n_uso_vivienda = case_when(
      is.na(n_locales_todos_estados) ~ NA_real_,
      n_locales_todos_estados == 0    ~ 0,
      TRUE                             ~ replace_na(n_uso_vivienda, 0)
    ),
    # computes the percentage of establishments used as housing for every case that doesn't have a 0 denominator, except sin_cobertura, which is 0
    pct_uso_vivienda = case_when(
      sin_cobertura                 ~ 0, 
      n_locales_todos_estados == 0  ~ NA_real_,
      TRUE                           ~ n_uso_vivienda / n_locales_todos_estados
    )
  ) %>%
  select(-sin_cobertura) %>%
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         n_locales_todos_estados, n_uso_vivienda, pct_uso_vivienda) %>%
  arrange(anio, distrito_num, barrio_num)



# join with the main table
censo_locales_barrio_anio <- censo_locales_barrio_anio %>%
  left_join(
    uso_vivienda_barrio_anio %>% select(anio, codigo_barrio, n_locales_todos_estados, n_uso_vivienda, pct_uso_vivienda), by = c("anio", "codigo_barrio"))
```

### Checks (02).

```{r compro 9}
# checks that no active premise is left without a recognized neighbourhood after reconstructing codigo_barrio
stopifnot(all(censo_locales_raw_todos$codigo_barrio %in% barrios_madrid$codigo_barrio))

# checks that the 8 economic groups appear as a column (none is left completely empty)
stopifnot(all(paste0("n_", cols_grupo) %in% names(conteo_grupo_barrio_anio)))

# checks that no neighbourhood is left unjoined
stopifnot(sum(is.na(censo_locales_barrio_anio$distrito_num)) == 0)

# checks the number of neighbourhoods per year:
conteo_barrios_censo <- censo_locales_barrio_anio %>% count(anio)
stopifnot(conteo_barrios_censo$n[conteo_barrios_censo$anio == 2017] == 128)
stopifnot(all(conteo_barrios_censo$n[conteo_barrios_censo$anio != 2017] == 131))

# checks the % are between 0-1 (NAs from a 0 denominator are ignored)
pct_cols <- paste0("pct_", cols_grupo)
stopifnot(all(sapply(censo_locales_barrio_anio[pct_cols], function(x) all(x >= 0 & x <= 1, na.rm = TRUE))))
```

### Saving

```{r save 9}
write_csv(censo_locales_barrio_anio, "../data/processed/censo_locales_barrio_anio.csv")
```

### View of the final table

```{r view 9}
#| eval: false
censo_locales_barrio_anio <- read_csv("../data/processed/censo_locales_barrio_anio.csv", show_col_types = FALSE)
View(censo_locales_barrio_anio)
```

## Source 10: residential churn (`churn_total/`)

**Dataset comments**:

These variables are built from 4 Banco de Datos files that record Padrón registrations and deregistrations. They are proxies for population turnover, resident inflow and outflow rates per neighbourhood-year. They are computed keeping the direction of the movement but collapsing 2 origins, giving 1) domestic registrations (altas) / deregistrations (bajas), internal to the city of Madrid, 2) residential registrations/deregistrations, external to the city. Dividing by the neighbourhood's total population to compute the rates normalizes the absolute counts by population size, so they are comparable between large and small neighbourhoods.

### Dataset cleaning

```{r limpieza 10}
parse_num_churn <- function(x) as.numeric(str_remove_all(x, "\\.")) # removes the thousands-separator period

leer_churn_archivo <- function(nombre_archivo) {
  archivo <- sprintf("../data/raw/churn_total/%s.xlsx", nombre_archivo)
  read_excel(archivo, sheet = 1, skip = 8, # sheet 1, skips 8 rows
             col_names = c("anio_txt", "distrito_txt", "barrio_txt", "total_txt", "espanola_txt", "extranjera_txt"), col_types = "text") %>% # forces them to be read as text
    filter(str_detect(anio_txt, "^\\d{4}$")) %>% # discards footer rows using the 4-digit year pattern

    filter(as.integer(anio_txt) <= 2023) %>% # filters years less than 2023
    filter(barrio_txt != distrito_txt) %>% # discards "Ciudad de Madrid" and the district-total rows by removing those rows that have the same text in both columns
    filter(str_detect(barrio_txt, "^\\d{3}\\.\\s")) %>% # only real neighbourhoods (3-digit pattern)
    filter(!str_detect(barrio_txt, regex("No consta", ignore_case = TRUE))) %>% # "000. No consta" does match the 3-digit pattern, so it is explicitly excluded

    transmute(
      anio = as.integer(anio_txt),
      codigo_barrio = str_extract(barrio_txt, "^\\d{3}"), # neighbourhood code extraction
      total = parse_num_churn(total_txt),
      espanola = parse_num_churn(espanola_txt), # applies the parse
      extranjera = parse_num_churn(extranjera_txt))
}


# total/española/extranjera are renamed with the flow's prefix, so the 4 can be joined without column overlap
renombrar_flujo_churn <- function(df, prefijo) {
  df %>% rename_with(~ paste0(prefijo, "_", .x), c(total, espanola, extranjera))
} # function that defines how the renaming is done

altas_internas_churn <- leer_churn_archivo("altas_domicilio_interno") %>% renombrar_flujo_churn("altas_internas") # prefix set for each file according to direction and origin
bajas_internas_churn <- leer_churn_archivo("bajas_domicilio_interno") %>% renombrar_flujo_churn("bajas_internas")
altas_externas_churn <- leer_churn_archivo("altas_residencia_externo") %>% renombrar_flujo_churn("altas_externas")
bajas_externas_churn <- leer_churn_archivo("bajas_residencia_externo") %>% renombrar_flujo_churn("bajas_externas")
```

#### Joining the 4 flows, totals and rates

```{r union 10}
# the 4 tables are joined into one, matching on anio and codigo_barrio
churn_join <- altas_internas_churn %>%
  full_join(bajas_internas_churn, by = c("anio", "codigo_barrio")) %>%
  full_join(altas_externas_churn, by = c("anio", "codigo_barrio")) %>%
  full_join(bajas_externas_churn, by = c("anio", "codigo_barrio"))

# computes total inflows and outflows (origin is collapsed, direction is kept)
churn_totales <- churn_join %>%
  mutate(
    entradas_totales_total = altas_internas_total + altas_externas_total,
    salidas_totales_total = bajas_internas_total + bajas_externas_total
  )

# uses the neighbourhood population already built in Source 6 (age bracket)
poblacion_para_churn <- tramo_edad_barrio_anio %>% select(anio, codigo_barrio, poblacion_total_barrio)

churn_con_poblacion <- churn_totales %>%
  left_join(poblacion_para_churn, by = c("anio", "codigo_barrio")) %>% # adds each neighbourhood-year's population from source 6
  left_join(barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio), by = "codigo_barrio") # adds barrios_madrid identifiers


# Computes 6 rates, dividing each flow and the totals by the neighbourhood-year population.
churn_total_barrio_anio <- churn_con_poblacion %>%
  mutate(
    tasa_altas_internas = altas_internas_total / poblacion_total_barrio,
    tasa_bajas_internas = bajas_internas_total / poblacion_total_barrio,
    tasa_altas_externas = altas_externas_total / poblacion_total_barrio,
    tasa_bajas_externas = bajas_externas_total / poblacion_total_barrio,
    tasa_entradas_totales = entradas_totales_total / poblacion_total_barrio,
    tasa_salidas_totales = salidas_totales_total / poblacion_total_barrio
  ) %>%

# the final columns are renamed
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,
         n_altas_internas = altas_internas_total, n_altas_internas_esp = altas_internas_espanola, n_altas_internas_ext = altas_internas_extranjera,
         n_bajas_internas = bajas_internas_total, n_bajas_internas_esp = bajas_internas_espanola, n_bajas_internas_ext = bajas_internas_extranjera,
         n_altas_externas = altas_externas_total, n_altas_externas_esp = altas_externas_espanola, n_altas_externas_ext = altas_externas_extranjera,
         n_bajas_externas = bajas_externas_total, n_bajas_externas_esp = bajas_externas_espanola, n_bajas_externas_ext = bajas_externas_extranjera,
         n_entradas_totales = entradas_totales_total, n_salidas_totales = salidas_totales_total,
         poblacion_total_barrio,
         tasa_altas_internas, tasa_bajas_internas, tasa_altas_externas, tasa_bajas_externas,
         tasa_entradas_totales, tasa_salidas_totales) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 10}
# checks the number of neighbourhoods in each year
stopifnot(all(count(churn_total_barrio_anio, anio)$n == 131))
stopifnot(sum(is.na(churn_total_barrio_anio$distrito_num)) == 0)

# checks that the rates are not negative
tasa_cols_churn <- c("tasa_altas_internas", "tasa_bajas_internas", "tasa_altas_externas",
                     "tasa_bajas_externas", "tasa_entradas_totales", "tasa_salidas_totales")
stopifnot(all(sapply(churn_total_barrio_anio[tasa_cols_churn], function(x) all(x >= 0, na.rm = TRUE))))

# checks the rows where the population left_join found no match: the expected ones are the 3 neighbourhoods of 2017 that tramo_edad/ does not cover.
barrios_sin_poblacion_churn <- churn_con_poblacion %>% filter(is.na(poblacion_total_barrio))
stopifnot(nrow(barrios_sin_poblacion_churn) == 3)
stopifnot(all(barrios_sin_poblacion_churn$anio == 2017))
stopifnot(all(sort(barrios_sin_poblacion_churn$codigo_barrio) == c("183", "193", "194")))

# checks internal consistency: city-wide total altas_domicilio_interno = city-wide total bajas_domicilio_interno

# 1. In the raw file
altas_internas_ciudad <- read_excel("../data/raw/churn_total/altas_domicilio_interno.xlsx", sheet = 1, skip = 8, col_names = c("anio_txt", "distrito_txt", "barrio_txt", "total_txt", "espanola_txt", "extranjera_txt"), col_types = "text") %>% # re-reads the raw domicile-registrations file
  filter(distrito_txt == "Ciudad de Madrid") %>% # selects the Ciudad de Madrid total
  transmute(anio = as.integer(anio_txt), total = parse_num_churn(total_txt))

bajas_internas_ciudad <- read_excel("../data/raw/churn_total/bajas_domicilio_interno.xlsx", sheet = 1, skip = 8, col_names = c("anio_txt", "distrito_txt", "barrio_txt", "total_txt", "espanola_txt", "extranjera_txt"), col_types = "text") %>% # re-reads the raw domicile-deregistrations file
  filter(distrito_txt == "Ciudad de Madrid") %>% # selects the Ciudad de Madrid total
  transmute(anio = as.integer(anio_txt), total = parse_num_churn(total_txt))

stopifnot(nrow(altas_internas_ciudad) == 7, nrow(bajas_internas_ciudad) == 7) # checks that 7 rows were included, one per year
stopifnot(all(altas_internas_ciudad$total == bajas_internas_ciudad$total)) # checks that the Ciudad de Madrid registration and deregistration totals in the raw files are equal


# 2. In the processed file versus the raw file
# checks consistency between the raw file's city total and the sum of your already-processed neighbourhoods (verifies a sound cleaning pipeline)
suma_altas_internas_procesado <- churn_total_barrio_anio %>% # registrations
  group_by(anio) %>%
  summarise(total = sum(n_altas_internas, na.rm = TRUE))

# known exception in 2023, the "Ciudad de Madrid" total in altas_domicilio_interno.xlsx does not match the sum of its own 21 districts (198349 vs 198342, a difference of -7) -> confirmed in the raw file
diferencia_altas <- suma_altas_internas_procesado$total - altas_internas_ciudad$total
stopifnot(all(diferencia_altas == if_else(suma_altas_internas_procesado$anio == 2023, -7, 0)))

suma_bajas_internas_procesado <- churn_total_barrio_anio %>%
  group_by(anio) %>%
  summarise(total = sum(n_bajas_internas, na.rm = TRUE))

# known exception in 2023, the "Ciudad de Madrid" total in bajas_domicilio_interno.xlsx does not match the sum of its own 21 districts (198349 vs 198003, a difference of -346) -> confirmed in the raw file
diferencia_bajas <- suma_bajas_internas_procesado$total - bajas_internas_ciudad$total
stopifnot(all(diferencia_bajas == if_else(suma_bajas_internas_procesado$anio == 2023, -346, 0)))
```

### Saving

```{r save 10}
write_csv(churn_total_barrio_anio, "../data/processed/churn_total_barrio_anio.csv")
```

### View of the final table

```{r view 10}
#| eval: false
churn_total_barrio_anio <- read_csv("../data/processed/churn_total_barrio_anio.csv", show_col_types = FALSE)
View(churn_total_barrio_anio)
```

## Source 11: municipal investment by district (`inversiones_ayto_distritos.xlsx`)

**Dataset comments**:

The variables extracted are investment budget and actual spending by district and the intensity of investment based on the budget and actual spending. The intensity is understood as the value divided by the district population. Since this data it only exists at district level, so each neighbourhood inherits its district's value, without adding real variance at neighbourhood level. The denominator of the investment-intensity variable is the district's population, which makes it comparable across districts of different sizes. The only population gap is the one in tramo_edad/ in 2017.

### Dataset cleaning

```{r limpieza 11}
# extracts the districts from barrios_madrid and applies the name normalization (defined earlier) to their names
distritos_norm_inv <- barrios_madrid %>%
  distinct(distrito_num, distrito_nombre) %>%
  mutate(distrito_key = normalizar_nombre(distrito_nombre))

inversiones_distrito_raw <- read_excel("../data/raw/inversiones_ayto_distritos.xlsx", sheet = "Sheet") %>%
  rename(anio = `Año`, id_area = `Id Área`, nombre_area = `Nombre Área`,
         presupuesto_gasto = `Presupuesto Gasto`, gasto_real = `Gasto Real`) %>%
  filter(anio >= 2017, anio <= 2023) %>% # filters to the study window
  filter(!id_area %in% c("NA", "NN")) %>% # discards "No distritalizable" and "Varios distritos"
  mutate(distrito_key = normalizar_nombre(nombre_area)) # normalizes the area = district name

inversiones_distrito <- inversiones_distrito_raw %>% # joins both tables
  left_join(distritos_norm_inv %>% select(distrito_key, distrito_num), by = "distrito_key")

```

#### District population and investment intensity

```{r tasas 11}
# reuses the table built in Source 6 that gives population per neighbourhood-year and aggregates it to district level
poblacion_distrito <- tramo_edad_barrio_anio %>%
  group_by(anio, distrito_num) %>%
  summarise(poblacion_total_distrito = sum(poblacion_total_barrio), .groups = "drop")

inversiones_distrito_pob <- inversiones_distrito %>%
  left_join(poblacion_distrito, by = c("anio", "distrito_num")) %>% # adds the district's population to the previous table
  mutate( # computes the two rates
    intensidad_inversion_presupuesto = presupuesto_gasto / poblacion_total_distrito, # budget
    intensidad_inversion_real = gasto_real / poblacion_total_distrito) # actual spending
```

#### Broadcast to neighbourhood

```{r broadcast 11}
barrios_sin_entidad_2017_inv <- c("183", "193", "194") # 2017's missing neighbourhoods

# removes those neighbourhoods only when the year is 2017, the rest stays the same.
grid_barrio_anio_inv <- map_dfr(2017:2023, function(anio) {
  base <- barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio)
  if (anio == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017_inv)
  base %>% mutate(anio = anio, .before = 1)
})

# broadcast: since several neighbourhoods share the same district, each one of them receives the same investment value from its district.
inversiones_ayto_distritos_barrio_anio <- grid_barrio_anio_inv %>%
  left_join(inversiones_distrito_pob %>%
              select(anio, distrito_num, # renamed with the _distrito suffix
                     presupuesto_gasto_distrito = presupuesto_gasto,
                     gasto_real_distrito = gasto_real,
                     poblacion_total_distrito,
                     intensidad_inversion_presupuesto_distrito = intensidad_inversion_presupuesto,
                     intensidad_inversion_real_distrito = intensidad_inversion_real),
            by = c("anio", "distrito_num")) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 11}
# checks that every district in both tables (distritos_norm_inv and inversiones_distrito_raw) found a match after the join
stopifnot(sum(is.na(inversiones_distrito$distrito_num)) == 0)

# checks that there are 21 neighbourhoods for each year in the resulting joined table
stopifnot(nrow(inversiones_distrito) == 21 * 7)

# checks that the broadcast left no neighbourhood without investment data (no NA)
stopifnot(sum(is.na(inversiones_ayto_distritos_barrio_anio)) == 0)

# every neighbourhood within the same district-year must have exactly the same value (correct broadcast) -> n=1
chk_broadcast_inv <- inversiones_ayto_distritos_barrio_anio %>%
  group_by(anio, distrito_num) %>% # groups by year and district
  summarise(n_valores_distintos = n_distinct(presupuesto_gasto_distrito), .groups = "drop") # counts how many distinct values there are in each group
stopifnot(all(chk_broadcast_inv$n_valores_distintos == 1))
```

### Saving

```{r save 11}
write_csv(inversiones_ayto_distritos_barrio_anio, "../data/processed/inversiones_ayto_distritos_barrio_anio.csv")
```

### View of the final table

```{r view 11}
#| eval: false
inversiones_ayto_distritos_barrio_anio <- read_csv("../data/processed/inversiones_ayto_distritos_barrio_anio.csv", show_col_types = FALSE)
View(inversiones_ayto_distritos_barrio_anio)
```

## Source 12: municipal investment by district and category (`inversiones_tipo_distrito/`)

**Dataset comments**:

The variables extracted in this section are the amount of investment (divided in budget and actual spending) of each type and its share compared to the total amount of budget and actual spending. This is a district-level source that complements Source 11 with the breakdown by spending category (for now all investment categories are extracted, even though not all of them appear every year in every district). Rows that have an empty `Gasto Real` - a category with a budget assigned but with the budget execution not reported - are kept as NA.

### Dataset cleaning

```{r limpieza 12}
# dictionary of names for the 15 investment categories
categorias_inversion <- tribble(
  ~id_linea, ~categoria,
  "X00", "caracter_general",
  "X01", "proteccion_civil_seguridad",
  "X02", "trafico_transporte",
  "X03", "proteccion_promocion_social",
  "X04", "salud_publica",
  "X05", "educacion",
  "X06", "urbanismo",
  "X07", "infraestructuras",
  "X08", "mantenimiento_urbano",
  "X09", "medio_ambiente",
  "X10", "cultura",
  "X11", "deportes_juventud_esparcimiento",
  "X12", "otros_bienes_sociales",
  "X13", "otros_bienes_economicos",
  "X14", "sectores_productivos"
)
categorias_nom <- categorias_inversion$categoria

archivos_inv_tipo <- list.files("../data/raw/inversiones_tipo_distrito", pattern = "^inv_tipo_.*\\.csv$", full.names = TRUE) # Looks within the folder for the file whose name exactly matches the pattern.
stopifnot(length(archivos_inv_tipo) == 21) # checks there are 21 files


# reads the files, function
leer_inv_tipo_archivo <- function(archivo) {
  distrito_nom <- str_remove(str_remove(basename(archivo), "^inv_tipo_"), "\\.csv$") # keeps the file path and strips out everything from the name except the district itself.
  distrito_key_archivo <- normalizar_nombre(str_replace_all(distrito_nom, "_", " ")) # replaces underscores with spaces and normalizes the district name

  read_csv(archivo, skip = 1, # skips the actual header and names the columns by hand
           col_names = c("anio", "id_linea", "nombre_linea", "inversion_detalle", "inversion_extra",
                         "presupuesto_gasto", "gasto_real"),
           col_types = cols(anio = col_integer(), id_linea = col_character(), nombre_linea = col_character(), inversion_detalle = col_character(), inversion_extra = col_character(), presupuesto_gasto = col_double(), gasto_real = col_double()), # defines the data types
           locale = locale(encoding = "UTF-8")) %>% # sets the encoding

    filter(is.na(inversion_detalle) & is.na(inversion_extra)) %>% # only uses the rows that have the investment details empty in the raw file (NA in R) because those are the ones that give each category's total figure
    filter(anio >= 2017, anio <= 2023) %>%
    transmute(distrito_key = distrito_key_archivo, anio, id_linea, presupuesto_gasto, gasto_real)
}


# applies the function to the 21 files and stacks the results into a single table
inv_tipo_todos <- map_dfr(archivos_inv_tipo, leer_inv_tipo_archivo)

# joins by normalized name with the district catalog, bringing in distrito_num
inv_tipo_join <- inv_tipo_todos %>%
  left_join(distritos_norm_inv %>% select(distrito_key, distrito_num), by = "distrito_key")

# checks that no file name was left without a district and that all 21 distinct districts appear
stopifnot(sum(is.na(inv_tipo_join$distrito_num)) == 0)
stopifnot(n_distinct(inv_tipo_join$distrito_key) == 21)

# joins each row with the dictionary of the 15 categories, bringing in the readable name (categoria)
inv_tipo_cat <- inv_tipo_join %>% left_join(categorias_inversion, by = "id_linea")
stopifnot(sum(is.na(inv_tipo_cat$categoria)) == 0) # every id_linea must correspond to one of the 15 categories
```

#### To wide format and computing the indicator

```{r ancho 12}
# converts to wide format, with each category in its own column/variable
inv_tipo_ancho <- inv_tipo_cat %>%
  select(anio, distrito_num, categoria, presupuesto_gasto, gasto_real) %>%
  pivot_wider(names_from = categoria, values_from = c(presupuesto_gasto, gasto_real), values_fill = 0)

# checks the number of rows
stopifnot(nrow(inv_tipo_ancho) == 21 * 7)
# checks the number of columns
stopifnot(ncol(inv_tipo_ancho) == 2 + 2 * length(categorias_nom))


# the denominator's total is Source 11's total investment (budgeted and actual)
inv_tipo_final_distrito <- inv_tipo_ancho %>%
  left_join(inversiones_distrito_pob %>% select(anio, distrito_num, presupuesto_gasto_distrito = presupuesto_gasto, gasto_real_distrito = gasto_real), by = c("anio", "distrito_num"))

# each mutate computes the % that each category represents of the total for its type (budget or actual spending)
inv_tipo_shares <- inv_tipo_final_distrito %>%
  mutate(across(all_of(paste0("presupuesto_gasto_", categorias_nom)), # every category of one type
                ~ .x / presupuesto_gasto_distrito,
                .names = "share_presupuesto_{sub('^presupuesto_gasto_', '', .col)}")) %>%
  mutate(across(all_of(paste0("gasto_real_", categorias_nom)), # ...and of the other
                ~ .x / gasto_real_distrito,
                .names = "share_gasto_real_{sub('^gasto_real_', '', .col)}"))
```

#### Broadcast to neighbourhood

```{r broadcast 12}
barrios_sin_entidad_2017_inv2 <- c("183", "193", "194")  # 2017's missing neighbourhoods

# removes those neighbourhoods only when the year is 2017, the rest stays the same.
grid_barrio_anio_inv2 <- map_dfr(2017:2023, function(anio) {
  base <- barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio)
  if (anio == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017_inv2)
  base %>% mutate(anio = anio, .before = 1)
})

# broadcast: since several neighbourhoods share the same district, each one of them receives the same investment-type % value from its district
inversiones_tipo_distrito_barrio_anio <- grid_barrio_anio_inv2 %>%
  left_join(inv_tipo_shares, by = c("anio", "distrito_num")) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r gasto real}
# the actual-spending columns can indeed have NA (budget execution not reported in that category)
cols_gasto_real <- inversiones_tipo_distrito_barrio_anio %>%
  select(starts_with("gasto_real"), starts_with("share_gasto_real"))
sum(is.na(cols_gasto_real)) # there are 676 cells with NA

# is it because they haven't been reported in the most recent years? no, it's a consistent pattern.
inversiones_tipo_distrito_barrio_anio %>%
  filter(if_any(starts_with("gasto_real"), is.na)) %>%
  distinct(anio, distrito_num) %>%
  count(anio)

# are there some investment types whose budget executions are reported less often? Yes, especially in infrastructure and transport.
inv_tipo_final_distrito %>%
  summarise(across(starts_with("gasto_real_"), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "categoria", values_to = "n_na") %>%
  arrange(desc(n_na))
```

```{r compro 12}
# checks NAs: no NA is expected in the budget columns - but they are expected in actual spending
cols_sin_na_esperado <- inversiones_tipo_distrito_barrio_anio %>%
  select(-starts_with("gasto_real"), -starts_with("share_gasto_real"))
stopifnot(sum(is.na(cols_sin_na_esperado)) == 0)

# checks the number of neighbourhoods per year
conteo_barrios_inv_tipo <- inversiones_tipo_distrito_barrio_anio %>% count(anio)
stopifnot(conteo_barrios_inv_tipo$n[conteo_barrios_inv_tipo$anio == 2017] == 128)
stopifnot(all(conteo_barrios_inv_tipo$n[conteo_barrios_inv_tipo$anio != 2017] == 131))

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(inversiones_tipo_distrito_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks that the budget percentages cover 100% of the total
suma_shares_presu <- inversiones_tipo_distrito_barrio_anio %>%
  mutate(suma = rowSums(across(all_of(paste0("share_presupuesto_", categorias_nom))))) %>%
  pull(suma)
stopifnot(all(abs(suma_shares_presu - 1) < 1e-6)) # == 1 with tolerance

# the actual-spending ones must also sum to 1, except in the district-years with some NA
gasto_real_cols_chk <- paste0("gasto_real_", categorias_nom)
suma_shares_real <- inversiones_tipo_distrito_barrio_anio %>%
  mutate(tiene_na = if_any(all_of(gasto_real_cols_chk), is.na),
         suma = rowSums(across(all_of(paste0("share_gasto_real_", categorias_nom))))) %>%
  filter(!tiene_na) %>%
  pull(suma)
stopifnot(all(abs(suma_shares_real - 1) < 1e-6)) # == 1 with tolerance
```

### Saving

```{r save 12}
write_csv(inversiones_tipo_distrito_barrio_anio, "../data/processed/inversiones_tipo_distrito_barrio_anio.csv")
```

### View of the final table

```{r view 12}
#| eval: false
inversiones_tipo_distrito_barrio_anio <- read_csv("../data/processed/inversiones_tipo_distrito_barrio_anio.csv", show_col_types = FALSE)
View(inversiones_tipo_distrito_barrio_anio)
```

## Source 13: nationality by country of origin (`nacionalidad_pais_distrito.xlsx`)

**Dataset comments**:

District-level source. Builds the predictor "percentage of population born in countries with high or very high HDI" (≥0.700, the two upper brackets of the UNDP classification). The denominator used for the calculation is the district's total population (not the migrant population). Spain is excluded from the numerator since, even though it has a high HDI, the indicator seeks to capture transnational gentrification. `indice_desarrollo_humano.csv` is used, computing the classification separately for each year from that year's HDI. The source's 200 individual countries are all covered, but 13 are left without an iso3 since they are not countries with their own UNDP HDI

### Dataset cleaning

```{r limpieza 13, message=FALSE}
raw_nac <- read_excel("../data/raw/nacionalidad_pais_distrito.xlsx", sheet = 1, col_names = FALSE) # no header is assumed

geografias_txt <- as.character(unlist(raw_nac[6, 4:ncol(raw_nac)])) # stores the text of row 6 from column 4 to the end of the file -> where the district names are.

# parsing function for cleaning
parse_num_nac <- function(x) {
  x <- if_else(x == "..", NA_character_, x) # statistical-confidentiality sentinel
  as.numeric(str_remove_all(x, "\\.")) # removes the thousands separator
}


datos_nac_largo <- bind_cols( # rebuilds the full table starting from row 8 (where the data starts)
    raw_nac[8:nrow(raw_nac), 1:3] %>% set_names(c("fecha_txt", "grupo", "pais")), # renames the first three columns
    raw_nac[8:nrow(raw_nac), 4:ncol(raw_nac)] %>% set_names(geografias_txt) # fills in the columns with the district names extracted earlier
  ) %>%
  mutate(fecha_txt = as.character(fecha_txt)) %>%
  filter(str_detect(fecha_txt, "^1 de enero de \\d{4}$")) %>% # discards footer rows since they don't match that pattern
  mutate(anio = as.integer(str_extract(fecha_txt, "\\d{4}"))) %>% # rewrites the date as a year (the number only)
  filter(anio >= 2017, anio <= 2023) %>% # filters to the study window
  select(-fecha_txt) %>%
  pivot_longer(cols = -c(anio, grupo, pais), names_to = "geografia_txt", values_to = "valor_txt") # applies pivot_longer to make grouping easier later on


datos_nac_distrito <- datos_nac_largo %>%
  filter(geografia_txt != "Ciudad de Madrid") %>% # removes the city-total row
  mutate(distrito_num = as.integer(str_extract(geografia_txt, "^\\d{2}")), # extracts the district code
         valor = parse_num_nac(valor_txt)) %>% # applies parse_num_nac
  select(anio, distrito_num, grupo, pais, valor)

# checks there are no NAs
stopifnot(sum(is.na(datos_nac_distrito$distrito_num)) == 0)
# checks that the number of districts is 21
stopifnot(n_distinct(datos_nac_distrito$distrito_num) == 21)


# keeps the row where grupo=total and pais=total, since that row holds the population totals across nationalities for each district
total_poblacion_nac <- datos_nac_distrito %>%
  filter(grupo == "Total", pais == "Total") %>%
  select(anio, distrito_num, poblacion_total_nac = valor)

paises_individuales_nac <- datos_nac_distrito %>% filter(grupo != pais) # excludes Total/España/Otro país/continent subtotals/Apátridas/No consta since those are the rows whose grupo-column text equals the pais-column text.
```

#### Classification of countries by HDI, year by year (`indice_desarrollo_humano.csv`)

```{r clasificacion 13}
idh_raw <- read_delim("../data/raw/indice_desarrollo_humano.csv", delim = ";",
                       locale = locale(encoding = "WINDOWS-1252"),
                       col_types = cols(.default = col_character()))

idh_alto_anio <- idh_raw %>%
  filter(!str_starts(iso3, "ZZ")) %>% # discards the aggregate rows (development level/region/world, iso3 with "ZZ" prefix)
  select(iso3, starts_with("hdi_20")) %>%  # keeps iso3 and the columns starting with "hdi_20"
  pivot_longer(starts_with("hdi_20"), names_to = "anio", names_prefix = "hdi_", values_to = "hdi_txt") %>% # switches to long format so that country-year is one observation (specifies the prefix of what will become the names)
  mutate(anio = as.integer(anio), hdi = suppressWarnings(as.numeric(hdi_txt))) %>% # converts year and hdi values to numeric
  filter(anio >= 2017, anio <= 2023) %>% # filters to the study window
  transmute(iso3, anio, hdi, es_alto = !is.na(hdi) & hdi >= 0.700) # If NA or <0.700 = FALSE / if >= 0.700 = TRUE -> boolean flag to build es_alto (high-HDI value using the official threshold)

stopifnot(n_distinct(idh_alto_anio$iso3) == 195) # checks there are 206 rows in the csv - 11 aggregates = 195 countries
stopifnot(nrow(idh_alto_anio) == 195 * 7) # checks there are 195 countries for each of the years


# crosswalk Spanish name (in nacionalidad_pais_distrito.xlsx) -> iso3 (indice_desarrollo_humano.csv).
crosswalk_pais_iso3 <- tribble(
  ~pais, ~iso3,
  "Afganistán", "AFG", "Albania", "ALB", "Alemania", "DEU", "Andorra", "AND",
  "Angola", "AGO", "Antigua y Barbuda", "ATG", "Arabia Saudí", "SAU", "Argelia", "DZA",
  "Argentina", "ARG", "Armenia", "ARM", "Australia", "AUS", "Austria", "AUT",
  "Azerbaiyán", "AZE", "Bahamas", "BHS", "Bahréin", "BHR", "Bangladesh", "BGD",
  "Barbados", "BRB", "Belarús", "BLR", "Belice", "BLZ", "Benín", "BEN",
  "Bhután", "BTN", "Bolivia", "BOL", "Bosnia y Herzegovina", "BIH", "Botswana", "BWA",
  "Brasil", "BRA", "Brunei", "BRN", "Bulgaria", "BGR", "Burkina Faso", "BFA",
  "Burundi", "BDI", "Bélgica", "BEL", "Cabo Verde", "CPV", "Camboya", "KHM",
  "Camerún", "CMR", "Canadá", "CAN", "Chad", "TCD", "Chile", "CHL",
  "China", "CHN", "Chipre", "CYP", "Colombia", "COL", "Comores", "COM",
  "Congo", "COG", "Corea", "KOR", "Corea del Norte", "PRK", "Costa Rica", "CRI",
  "Costa de Marfil", "CIV", "Croacia", "HRV", "Cuba", "CUB", "Dinamarca", "DNK",
  "Djibouti", "DJI", "Dominica", "DMA", "Ecuador", "ECU", "Egipto", "EGY",
  "El Salvador", "SLV", "Emiratos Árabes Unidos", "ARE", "Eritrea", "ERI", "Eslovenia", "SVN",
  "Estados Unidos de América", "USA", "Estonia", "EST", "Etiopía", "ETH", "Fiji", "FJI",
  "Filipinas", "PHL", "Finlandia", "FIN", "Francia", "FRA", "Gabón", "GAB",
  "Gambia", "GMB", "Georgia", "GEO", "Ghana", "GHA", "Granada", "GRD",
  "Grecia", "GRC", "Guatemala", "GTM", "Guinea", "GIN", "Guinea Ecuatorial", "GNQ",
  "Guinea-Bissau", "GNB", "Guyana", "GUY", "Haití", "HTI", "Honduras", "HND",
  "Hungría", "HUN", "India", "IND", "Indonesia", "IDN", "Iraq", "IRQ",
  "Irlanda", "IRL", "Irán", "IRN", "Islandia", "ISL", "Islas Marshall", "MHL",
  "Israel", "ISR", "Italia", "ITA", "Jamaica", "JAM", "Japón", "JPN",
  "Jordania", "JOR", "Kazajstán", "KAZ", "Kenia", "KEN", "Kirguistán", "KGZ",
  "Kuwait", "KWT", "Laos", "LAO", "Lesotho", "LSO", "Letonia", "LVA",
  "Liberia", "LBR", "Libia", "LBY", "Liechtenstein", "LIE", "Lituania", "LTU",
  "Luxemburgo", "LUX", "Líbano", "LBN", "Macedonia del Norte", "MKD", "Madagascar", "MDG",
  "Malasia", "MYS", "Malawi", "MWI", "Maldivas", "MDV", "Mali", "MLI",
  "Malta", "MLT", "Marruecos", "MAR", "Mauricio", "MUS", "Mauritania", "MRT",
  "Micronesia", "FSM", "Moldavia", "MDA", "Mongolia", "MNG", "Montenegro", "MNE",
  "Mozambique", "MOZ", "Myanmar", "MMR", "México", "MEX", "Mónaco", "MCO",
  "Namibia", "NAM", "Nepal", "NPL", "Nicaragua", "NIC", "Nigeria", "NGA",
  "Noruega", "NOR", "Nueva Zelanda", "NZL", "Níger", "NER", "Omán", "OMN",
  "Pakistán", "PAK", "Palestina", "PSE", "Panamá", "PAN", "Papúa Nueva Guinea", "PNG",
  "Paraguay", "PRY", "Países Bajos", "NLD", "Perú", "PER", "Polonia", "POL",
  "Portugal", "PRT", "Qatar", "QAT", "Reino Unido", "GBR", "República Centroafricana", "CAF",
  "República Checa", "CZE", "República Democrática del Congo", "COD", "República Dominicana", "DOM", "República Eslovaca", "SVK", "Ruanda", "RWA", "Rumanía", "ROU", "Rusia", "RUS",
  "Samoa", "WSM", "San Cristóbal y Nieves", "KNA", "San Marino", "SMR",
  "San Vicente y las Granadinas", "VCT", "Santa Lucía", "LCA",
  "Santo Tomé y Príncipe", "STP", "Senegal", "SEN", "Serbia", "SRB", "Seychelles", "SYC",
  "Sierra Leona", "SLE", "Singapur", "SGP", "Siria", "SYR", "Somalia", "SOM",
  "Sri Lanka", "LKA", "Sudáfrica", "ZAF", "Sudán", "SDN", "Sudán del Sur", "SSD",
  "Suecia", "SWE", "Suiza", "CHE", "Surinam", "SUR", "Swazilandia", "SWZ",
  "Tailandia", "THA", "Tanzania", "TZA", "Tayikistán", "TJK", "Timor Oriental", "TLS",
  "Togo", "TGO", "Trinidad y Tobago", "TTO", "Turkmenistán", "TKM", "Turquía", "TUR",
  "Túnez", "TUN", "Ucrania", "UKR", "Uganda", "UGA", "Uruguay", "URY",
  "Uzbekistán", "UZB", "Vanuatu", "VUT", "Venezuela", "VEN", "Vietnam", "VNM",
  "Yemen", "YEM", "Zambia", "ZMB", "Zimbabwe", "ZWE"
)
```

### Broadcast to neighbourhood

```{r broadcast 13}
poblacion_idh_alto <- paises_individuales_nac %>%
  left_join(crosswalk_pais_iso3, by = "pais") %>% # joins paises_individuales_nac rows with their iso3
  left_join(idh_alto_anio %>% select(iso3, anio, es_alto), by = c("iso3", "anio")) %>% # also joins them with their HDI classification for that specific year
  filter(es_alto) %>% # discards countries that don't have a high HDI
  group_by(anio, distrito_num) %>%
  summarise(poblacion_idh_alto = sum(valor, na.rm = TRUE), .groups = "drop") # sums the population (ignoring NA).

# this left_join produces NA in case a district-year has no country with high HDI; when it appears in poblacion_idh_alto it should act as 0 (no population)
nacionalidad_idh_distrito <- total_poblacion_nac %>%
  left_join(poblacion_idh_alto, by = c("anio", "distrito_num")) %>%
  mutate(poblacion_idh_alto = replace_na(poblacion_idh_alto, 0),
         pct_poblacion_idh_alto = poblacion_idh_alto / poblacion_total_nac) # computes the indicator



barrios_sin_entidad_2017_nac <- c("183", "193", "194") # 2017's missing neighbourhoods
# removes those neighbourhoods only when the year is 2017, the rest stays the same.
grid_barrio_anio_nac <- map_dfr(2017:2023, function(anio) {
  base <- barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio)
  if (anio == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017_nac)
  base %>% mutate(anio = anio, .before = 1)
})

# broadcast: since several neighbourhoods share the same district, each one of them receives the same % of high-HDI population value
nacionalidad_pais_distrito_barrio_anio <- grid_barrio_anio_nac %>%
  left_join(nacionalidad_idh_distrito %>% select(anio, distrito_num, # renaming
                                                   poblacion_total_distrito_nac = poblacion_total_nac,
                                                   poblacion_idh_alto_distrito = poblacion_idh_alto,
                                                   pct_poblacion_idh_alto_distrito = pct_poblacion_idh_alto),
            by = c("anio", "distrito_num")) %>% arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 13}
# crosswalk checks:
stopifnot(nrow(crosswalk_pais_iso3) == 187) # checks there are 187 countries
stopifnot(sum(duplicated(crosswalk_pais_iso3$pais)) == 0) # checks there are no duplicate countries
stopifnot(sum(duplicated(crosswalk_pais_iso3$iso3)) == 0) # checks there is no duplicate iso3
stopifnot(!"España" %in% crosswalk_pais_iso3$pais) # checks that Spain has been excluded
stopifnot(all(crosswalk_pais_iso3$pais %in% unique(paises_individuales_nac$pais))) # catches name typos in the Spanish names
stopifnot(all(crosswalk_pais_iso3$iso3 %in% unique(idh_alto_anio$iso3))) # catches typos in the iso3 codes

paises_sin_iso3_esperados <- c(
  "América Central y Caribe", "Antiguos territorios soberanía española", "Antillas Neerlandesas",
  "Aruba", "Islas Vírgenes", "Otros países de Europa", "Otros países o territorios de América del Sur",
  "Otros países o territorios de Asia", "Otros países o territorios de Oceanía",
  "Otros países o territorios de África", "Puerto Rico", "Taiwán", "Yugoslavia")
stopifnot(setequal(setdiff(unique(paises_individuales_nac$pais), crosswalk_pais_iso3$pais), paises_sin_iso3_esperados)) # confirms that the 13 countries without iso3 are the expected ones


# checks that the year-specific threshold does capture real movement and not just year-to-year redundancy
paises_frontera_chk <- crosswalk_pais_iso3 %>%
  left_join(idh_alto_anio, by = "iso3") %>%
  filter(!is.na(es_alto)) %>%
  group_by(pais) %>%
  filter(n_distinct(es_alto) > 1) %>%
  ungroup()
stopifnot(n_distinct(paises_frontera_chk$pais) > 0)

# correct broadcast: every neighbourhood within the same district-year shares the same value
chk_broadcast_nac <- nacionalidad_pais_distrito_barrio_anio %>%
  group_by(anio, distrito_num) %>%
  summarise(n_valores_distintos = n_distinct(pct_poblacion_idh_alto_distrito), .groups = "drop")
stopifnot(all(chk_broadcast_nac$n_valores_distintos == 1))


# checks that the indicator falls within the 0-1 range
stopifnot(all(nacionalidad_pais_distrito_barrio_anio$pct_poblacion_idh_alto_distrito >= 0 &
              nacionalidad_pais_distrito_barrio_anio$pct_poblacion_idh_alto_distrito <= 1))

# checks there are no NAs
stopifnot(sum(is.na(nacionalidad_pais_distrito_barrio_anio)) == 0)

# checks the number of neighbourhoods in each year
conteo_barrios_nac <- nacionalidad_pais_distrito_barrio_anio %>% count(anio)
stopifnot(conteo_barrios_nac$n[conteo_barrios_nac$anio == 2017] == 128)
stopifnot(all(conteo_barrios_nac$n[conteo_barrios_nac$anio != 2017] == 131))
```

### Saving

```{r save 13}
write_csv(nacionalidad_pais_distrito_barrio_anio, "../data/processed/nacionalidad_pais_distrito_barrio_anio.csv")
```

### View of the final table

```{r view 13}
#| eval: false
nacionalidad_pais_distrito_barrio_anio <- read_csv("../data/processed/nacionalidad_pais_distrito_barrio_anio.csv", show_col_types = FALSE)
View(nacionalidad_pais_distrito_barrio_anio)
```

## Source 14: green space by district (`zonas_verdes_distrito/`)

**Dataset comments**:

District-level source. Uses the surface area of "significant parks and green spaces" (sheet `ZV Y PARQUES SIGNIFIC. DISTRITO`), - motivated by De la Cruz & Ortín (2025). A normalized variable `zonas_verdes_significativas_m2_per_capita` (significant m² / district population) is computed.

### Dataset cleaning

```{r limpieza 14, message=FALSE}
# reading per year
leer_zonas_verdes_significativas_anio <- function(anio) {
  archivo <- sprintf("../data/raw/zonas_verdes_distrito/zonas_verdes_%d.xlsx", anio)
  raw <- read_excel(archivo, sheet = "ZV Y PARQUES SIGNIFIC. DISTRITO")

  col_distrito <- names(raw)[str_detect(names(raw), "^DISTRITO$")] # looks for the column matching distrito
  col_ha <- names(raw)[str_detect(names(raw), regex("\\bha\\b", ignore_case = TRUE))] # looks for the column containing Ha: the surface area
  stopifnot(length(col_distrito) == 1, length(col_ha) == 1) # verifies that each search found one column

  raw %>% # table with two variables: district name and park hectares
    transmute(distrito_txt = .data[[col_distrito]],
              ha_parque = as.numeric(.data[[col_ha]])) %>%
    filter(!is.na(distrito_txt), !is.na(ha_parque)) %>% # keeps only the ones that do have values in both columns
    filter(!str_detect(distrito_txt, regex("total", ignore_case = TRUE))) %>% # removes total rows
    group_by(distrito_txt) %>% # groups by district... to sum all the hectares
    summarise(zonas_verdes_significativas_ha = sum(ha_parque), .groups = "drop") %>%
    mutate(anio = anio, .before = 1) # adds the year column
}

# applies the function to the 7 years and stacks the results into a single table
zonas_verdes_significativas_todos <- map_dfr(2017:2023, leer_zonas_verdes_significativas_anio)

zonas_verdes_significativas_join <- zonas_verdes_significativas_todos %>%
  mutate(distrito_key = normalizar_nombre(distrito_txt)) %>% # normalizes the district names (function already defined earlier)
  left_join(barrios_madrid %>% distinct(distrito_num, distrito_nombre) %>%
              mutate(distrito_key = normalizar_nombre(distrito_nombre)) %>% select(distrito_key, distrito_num),
            by = "distrito_key") # joins with barrios_madrid to bring in district identifiers
```

#### Normalization: m² per inhabitant

```{r per capita 14}
# computes the district population per year by summing all its neighbourhoods' population (uses tramo_edad_barrio_anio) -> denominator
poblacion_distrito_zv <- tramo_edad_barrio_anio %>%
  group_by(anio, distrito_num) %>%
  summarise(poblacion_total_distrito_zv = sum(poblacion_total_barrio), .groups = "drop")

# indicator computation
zonas_verdes_distrito_final <- zonas_verdes_significativas_join %>%
  select(anio, distrito_num, zonas_verdes_significativas_ha) %>%
  left_join(poblacion_distrito_zv, by = c("anio", "distrito_num")) %>% # joins the significant-park surface area with the district population
  mutate(zonas_verdes_significativas_m2_per_capita = zonas_verdes_significativas_ha * 10000 / poblacion_total_distrito_zv) # computation of m² per inhabitant / * 10000 converts hectares to m²
```

#### Broadcast to neighbourhood

```{r broadcast 14}
barrios_sin_entidad_2017_zv <- c("183", "193", "194") # 2017's missing neighbourhoods
# removes those neighbourhoods only when the year is 2017, the rest stays the same.
grid_barrio_anio_zv <- map_dfr(2017:2023, function(anio) {
  base <- barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio)
  if (anio == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017_zv)
  base %>% mutate(anio = anio, .before = 1)
})

# broadcast: since several neighbourhoods share the same district, each one of them receives the same significant-green-space-per-capita value
zonas_verdes_distrito_barrio_anio <- grid_barrio_anio_zv %>%
  left_join(zonas_verdes_distrito_final %>%
              select(anio, distrito_num,
                     zonas_verdes_significativas_ha_distrito = zonas_verdes_significativas_ha,
                     zonas_verdes_significativas_m2_per_capita_distrito = zonas_verdes_significativas_m2_per_capita), by = c("anio", "distrito_num")) %>%
  arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 14}
# checks that no neighbourhood was left without a match when joining the table with (barrios_madrid)
stopifnot(sum(is.na(zonas_verdes_significativas_join$distrito_num)) == 0)

# checks that all 21 districts appear each year after joining the table with (barrios_madrid)
stopifnot(nrow(zonas_verdes_significativas_join) == 21 * 7)

# checks that no neighbourhood was left without a match when joining the zonas_verdes_significativas_join and poblacion_distrito_zv tables
stopifnot(sum(is.na(zonas_verdes_distrito_final)) == 0)

# checks there are no NAs in the final table
stopifnot(sum(is.na(zonas_verdes_distrito_barrio_anio)) == 0)

# checks that the indicator is never negative
stopifnot(all(zonas_verdes_distrito_barrio_anio$zonas_verdes_significativas_m2_per_capita_distrito >= 0))

# checks the number of neighbourhoods for each year
conteo_barrios_zv_sig <- zonas_verdes_distrito_barrio_anio %>% count(anio)
stopifnot(conteo_barrios_zv_sig$n[conteo_barrios_zv_sig$anio == 2017] == 128)
stopifnot(all(conteo_barrios_zv_sig$n[conteo_barrios_zv_sig$anio != 2017] == 131))

# correct broadcast: every neighbourhood within the same district-year shares the same value
chk_broadcast_zv_sig <- zonas_verdes_distrito_barrio_anio %>%
  group_by(anio, distrito_num) %>%
  summarise(n_valores_distintos = n_distinct(zonas_verdes_significativas_ha_distrito), .groups = "drop")
stopifnot(all(chk_broadcast_zv_sig$n_valores_distintos == 1))
```

### Saving

```{r save 14}
write_csv(zonas_verdes_distrito_barrio_anio, "../data/processed/zonas_verdes_distrito_barrio_anio.csv")
```

### View of the final table

```{r view 14}
#| eval: false
zonas_verdes_distrito_barrio_anio <- read_csv("../data/processed/zonas_verdes_distrito_barrio_anio.csv", show_col_types = FALSE)
View(zonas_verdes_distrito_barrio_anio)
```

## Source 15: distance to metro (`distancia_metro/`)

**Dataset comments**:

There are two files: `estaciones_metro_nodes.csv` (holds the stations with X/Y coordinates and their attributes), `itinerarios_metro_edges.csv` (holds the itineraries, the ordered sequence of stations and the distance to the previous station).

This variable is at neighbourhood level but has no year dimension since it is the current snapshot of the CRTM metro network. Using the current snapshot does not introduce much variation since the only modification completed and opened within the city of Madrid's limits between 2017 and 2026 was the creation of the Arroyofresno station (Line 7) in 2019. In the raw data there is an error in the Arroyofresno station's start date, which actually corresponds to April 2019, this being the only relevant metro-network modification. It will be accounted for through a manual correction in the years 2017 and 2018.

For every neighbourhood with a station of its own, the built variable represents the minimum number of segments (track segments between two consecutive stations of the same itinerary in a given direction) that must be traveled on the shortest path (measured in number of segments) from that station to Sol, treated as the reference "center" since it is kilometer zero of the Spanish radial highways. It is computed by building the metro network as a graph (stations = nodes; segments between consecutive stations = edges with weight 1 each, a transfer between lines at the same physical station = edge with weight 0, not counted as a traveled segment) and applying Dijkstra (via `igraph::distances()`) from Sol to the rest of the stations. There are 32 neighbourhoods that have no metro station of their own, so they are left with `paradas_red_centro` as NA.

### Reading and building the graph

```{r lectura 15, message=FALSE}
# reading
estaciones_raw <- read_csv("../data/raw/distancia_metro/estaciones_metro_nodes.csv",
                            col_types = cols(.default = col_character()),
                            locale = locale(encoding = "UTF-8"))

itinerarios_raw <- read_csv("../data/raw/distancia_metro/itinerarios_metro_edges.csv",
                             col_types = cols(.default = col_character()),
                             locale = locale(encoding = "UTF-8"))
```

```{r nodos 15}
# nodes
estaciones <- estaciones_raw %>%
  transmute(
    codigo_estacion = CODIGOESTACION, # renaming
    denominacion = DENOMINACION,
    nombre_norm = str_squish(stringi::stri_trans_general(str_to_upper(DENOMINACION), "Latin-ASCII")), # applies space cleanup, normalizes everything to uppercase, without accents, etc.
    municipio = CODIGOMUNICIPIO,
    distrito_metro = DISTRITO,
    barrio_metro = BARRIO,
    fecha_inicio = suppressWarnings(as.integer(str_sub(FECHAINICIO, 1, 4))), # keeps only the first 4 characters of the start date
    x = suppressWarnings(as.numeric(X)), # stores the coordinates as numeric
    y = suppressWarnings(as.numeric(Y)))
```

```{r aristas de red 15}
# edges
itinerarios <- itinerarios_raw %>%
  transmute(codigo_itinerario = CODIGOITINERARIO, numero_orden = as.integer(NUMEROORDEN), codigo_estacion = CODIGOESTACION) %>%
  arrange(codigo_itinerario, numero_orden) # sorts the table by itinerary and then by stop order within that itinerary (each itinerary is a metro line traveled in one direction with its stops in order).

# creates the edges between two consecutive stations
tramos_red <- itinerarios %>%
  group_by(codigo_itinerario) %>% # groups by line-direction = itinerary
  mutate(codigo_estacion_previa = lag(codigo_estacion)) %>% # lag is a function that shifts a vector one position back, so it gives, in each row, the previous row's station within that same group. Each itinerary's first station has no "previous", so it will get NA.
  ungroup() %>%
  filter(!is.na(codigo_estacion_previa)) %>% # discards each itinerary's first stop, since it has NA, as an edge cannot be created without an origin node.
  transmute(desde = codigo_estacion_previa, hasta = codigo_estacion, weight = 1) # assigned weight = 1
```

Interchange stations have a different CODIGOESTACION for each line passing through there. For the graph not to be split into per-line components, they need to be grouped.

```{r aristas de transbordo 15, warning=FALSE}
# transfers are detected by grouping by repeated normalized name:
grupos_nombre_metro <- estaciones %>%
  filter(!is.na(denominacion), denominacion != "", !is.na(x), !is.na(y)) %>% # filters out the ones without a name or coordinates
  add_count(nombre_norm, name = "n_estaciones_nombre") %>% # add_count: a dplyr shortcut that does group_by(nombre_norm) + row count per group + left_join() to the original table + ungroup() / the result is a new column with how many times that nombre_norm repeats across the whole table.
  filter(n_estaciones_nombre > 1) %>% # keeps the ones that repeat more than once
  select(nombre_norm, codigo_estacion, x, y)



# a self-join is done, joining grupos_nombre_metro with grupos_nombre_metro, in order to get every transfer combination
# Example with Sol: 3 rows with code S1, S2, S3. The join generates 3×3 = 9 combinations for that name (S1,S1), (S1,S2), (S1,S3), (S2,S1), (S2,S2), (S2,S3), (S3,S1), (S3,S2), (S3,S3)
transbordos_pares <- grupos_nombre_metro %>%
  inner_join(grupos_nombre_metro, by = "nombre_norm", suffix = c("_a", "_b")) %>% # since both copies of the table have the same columns, the _a and _b suffixes are added so as not to have two columns with the same name.
  filter(codigo_estacion_a < codigo_estacion_b) %>% # avoids duplicate pairs (S1,S3) vs. (S3,S1) and self-pairs (S3,S3) / it doesn't really evaluate < in a meaningful way but the condition works since it (arbitrarily) picks one of the two directions.
  mutate(distancia_m = sqrt((x_a - x_b)^2 + (y_a - y_b)^2)) # for each of those pairs computes the Euclidean distance between the two stations by directly applying the Pythagorean theorem to the coordinates


# builds a table with a weight-0 column, transfer origin and destination station
transbordos_edges <- transbordos_pares %>% transmute(desde = codigo_estacion_a, hasta = codigo_estacion_b, weight = 0)

# stacks the two edge tables vertically (one below the other) with their desde/hasta/weight columns
todos_los_tramos <- bind_rows(tramos_red, transbordos_edges)


# from the full station table, the ones with a known start date at or before 2023 are kept -> in practice this modifies almost no station.
estaciones_activas <- estaciones %>% filter(!is.na(fecha_inicio), fecha_inicio <= 2023)

# discards any edge that connects to a station excluded by the previous filter
tramos_activos <- todos_los_tramos %>%
  filter(desde %in% estaciones_activas$codigo_estacion, hasta %in% estaciones_activas$codigo_estacion)
```

```{r grafo 15}
# builds the GRAPH object
grafo_metro <- graph_from_data_frame(tramos_activos, # edges table
                                     directed = FALSE, # non-directional network
                                     vertices = estaciones_activas %>% # nodes table
               select(codigo_estacion)) 
```

### Shortest path from Sol

```{r desde sol 15}
# creates the node that will be used as the center reference

# keeps only the rows whose normalized name is exactly "SOL"
sol_codigos <- estaciones_activas %>% filter(nombre_norm == "SOL") %>% pull(codigo_estacion)
stopifnot(length(sol_codigos) == 3) # Sol has 3 codigoestacion values (lines 1, 2 and 3)


# creates the "CENTRO_HUB" node connected with weight 0 to Sol's 3 variants, in order to be able to ask igraph::distances() for a single origin
grafo_con_hub <- grafo_metro %>% # starts from the general graph
  add_vertices(1, name = "CENTRO_HUB") %>% # adds 1 node named CENTRO_HUB
  add_edges(c("CENTRO_HUB", sol_codigos[1], "CENTRO_HUB", sol_codigos[2], "CENTRO_HUB", sol_codigos[3]),
            weight = 0) # origin-destination vector (x3), each edge is added with weight 0 since it is a transfer


# computes the distance value from the center -> distances() computes the shortest-path distances from vertex v to every other vertex in the graph, using the weight vector.
paradas_desde_centro <- distances(grafo_con_hub, v = "CENTRO_HUB", weights = E(grafo_con_hub)$weight)[1, ]
# E(grafo_con_hub) is an igraph function that returns the graph's edge sequence, and $weight extracts the weight attribute for each of them
# [1, ] selects the matrix's first column (the only one there is) -> the 1st row starts from v and each column is one of the graph's vertices.


paradas_tbl <- tibble(codigo_estacion = names(paradas_desde_centro),
                      paradas_desde_centro) %>% # names is actually taken from the previous vector extracted with [1, ] which, coming from a matrix, keeps the original stop names as the names attribute.
  filter(codigo_estacion != "CENTRO_HUB") %>% # removed since it is not a real station itself
  mutate(paradas_desde_centro = if_else(is.infinite(paradas_desde_centro), NA_real_, paradas_desde_centro))
# Inf is what distances() returns for any vertex that cannot be reached from the origin -> in that case it is replaced with NA, and otherwise the same value is kept.

# checks that every active station is reachable from Sol
stopifnot(sum(is.na(paradas_tbl$paradas_desde_centro)) == 0)
```

### Aggregation to neighbourhood and broadcast to the study years

```{r neighbourhood 15}
estaciones_madrid <- estaciones_activas %>% # adds each active station's paradas_desde_centro value
  left_join(paradas_tbl, by = "codigo_estacion") %>%
  filter(municipio == "079" & !is.na(barrio_metro) & barrio_metro != "") # filters to only the ones in the city of Madrid (Madrid municipality) AND with a neighbourhood that is neither NA nor empty

# computes the distance to the neighbourhood -> picking the shortest path for each of them
distancia_barrio <- estaciones_madrid %>%
  rename(codigo_barrio = barrio_metro) %>%
  group_by(codigo_barrio) %>% # groups by the new codigo_barrio name
  summarise(paradas_red_centro = min(paradas_desde_centro), n_estaciones_metro = n(), .groups = "drop")
  # paradas_red_centro: minimum number of stops along the shortest path to Sol
  # n_estaciones_metro: how many metro stations that neighbourhood has within its boundaries

# neighbourhoods without any station of their own: NA in stops, 0 in number of stations
distancia_metro_barrio <- barrios_madrid %>% # starts from barrios_madrid and joins distancia_barrio to it -> this way all 131 are kept, and neighbourhoods with no match in distancia_barrio are left with NA
  left_join(distancia_barrio, by = "codigo_barrio") %>%
  mutate(n_estaciones_metro = replace_na(n_estaciones_metro, 0)) # replaces NA with 0 in n_estaciones_metro
```

```{r broadcast 15}
barrios_sin_entidad_2017_metro <- c("183", "193", "194") # 2017's missing neighbourhoods

# removes those neighbourhoods only when the year is 2017, the rest stays the same.
# runs the function for each year and stacks the results
distancia_metro_barrio_anio <- map_dfr(2017:2023, function(anio_actual) {
  base <- distancia_metro_barrio
  if (anio_actual == 2017) base <- base %>% filter(!codigo_barrio %in% barrios_sin_entidad_2017_metro)
  base %>% mutate(anio = anio_actual, .before = 1)
})
```

### Manual correction of one neighbourhood

Arroyofresno is in the neighbourhood of Mirasierra with codigo_barrio "087". This neighbourhood has 3 other stations of its own (Pitis, Mirasierra, Paco de Lucía), so n_estaciones_metro should be 3 (not 4) in 2017 and 2018, and 4 from 2019 onward. As for paradas_red_centro, this column is not corrected since the minimum number of stops to Sol in that neighbourhood is determined by the Mirasierra station.

```{r correccion 15}
# investigates which station determines the minimum number of stops in the Mirasierra neighbourhood (087)
paradas_mirasierra <- estaciones_madrid %>%
  filter(barrio_metro == "087") %>%
  select(codigo_estacion, denominacion, paradas_desde_centro) %>%
  arrange(paradas_desde_centro)
paradas_mirasierra

minimo_con_arroyofresno <- min(paradas_mirasierra$paradas_desde_centro)
minimo_sin_arroyofresno <- paradas_mirasierra %>%
  filter(denominacion != "ARROYOFRESNO") %>%
  pull(paradas_desde_centro) %>%
  min()

# confirms that the minimum doesn't change if Arroyofresno is removed
stopifnot(minimo_con_arroyofresno == minimo_sin_arroyofresno)


# Manual correction for Arroyofresno
distancia_metro_barrio_anio <- distancia_metro_barrio_anio %>%
  mutate(n_estaciones_metro = if_else(codigo_barrio == "087" & anio %in% c(2017, 2018), n_estaciones_metro - 1, n_estaciones_metro)) # if code 087 and years 2017-2018 hold, 1 station is subtracted, otherwise the same value is kept.
```

### Checks (02).

```{r compro 15}
# checks there is only 1 row per station in estaciones_raw (no duplicates)
stopifnot(n_distinct(estaciones_raw$CODIGOESTACION) == nrow(estaciones_raw))

# checks there are no NAs in tramos_red (edges)
stopifnot(sum(is.na(tramos_red$weight)) == 0)

# checks the number of edges for each itinerary (if an itinerary has n stops, it generates n−1 edges)
stopifnot(nrow(tramos_red) == nrow(itinerarios_raw) - n_distinct(itinerarios_raw$CODIGOITINERARIO))

# checks that at least some transfer was detected
stopifnot(nrow(transbordos_pares) > 0)

# checks the transfers with the greatest distance between their stations -> consistent with the Madrid metro.
transbordos_pares %>% slice_max(distancia_m, n = 5) %>%
  select(nombre_norm, codigo_estacion_a, codigo_estacion_b, distancia_m)

# checks that there are 99 neighbourhoods with a station of their own
stopifnot(sum(!is.na(distancia_metro_barrio$paradas_red_centro)) == 99)

# checks that the distance from Sol to itself is 0
stopifnot(distancia_metro_barrio$paradas_red_centro[distancia_metro_barrio$codigo_barrio == "016"] == 0)

# checks the number of neighbourhoods per year
conteo_barrios_metro <- distancia_metro_barrio_anio %>% count(anio)
stopifnot(conteo_barrios_metro$n[conteo_barrios_metro$anio == 2017] == 128)
stopifnot(all(conteo_barrios_metro$n[conteo_barrios_metro$anio != 2017] == 131))
```

### Saving

```{r save 15}
write_csv(distancia_metro_barrio_anio, "../data/processed/distancia_metro_barrio_anio.csv")
```

```{r saving para script 02}
# This intermediate table is saved so it can be reused in 02
estaciones_paradas_export <- estaciones_activas %>%
  left_join(paradas_tbl, by = "codigo_estacion") %>%
  filter(!is.na(x), !is.na(y)) %>%
  select(codigo_estacion, municipio, distrito_metro, barrio_metro, x, y, paradas_desde_centro)

write_csv(estaciones_paradas_export, "../data/processed/estaciones_metro_paradas.csv")
```

### View of the final table

```{r view 15}
#| eval: false
distancia_metro_barrio_anio <- read_csv("../data/processed/distancia_metro_barrio_anio.csv", show_col_types = FALSE)
View(distancia_metro_barrio_anio)
```

## Source 16: Distance to the center (`geometria_barrios_extracted/`)

**Dataset comments**:

Two SHP files with the polygon of each Madrid neighbourhood, at two different dates: `data/raw/geometria_barrios_extracted/Barrios2017/Neighbourhoods.shp` and `data/raw/geometria_barrios_extracted/Barrios2021/Barrios_20210712.shp`. The variable doesn't offer much variance across years; nonetheless, 2017 and 2021 data (dates available from the Ayto. de Madrid data) were used to capture the administrative changes that affect a neighbourhood's shape and centroid location (Timón, Casco Histórico de Vicálvaro, Valdefuentes, Palomas and Alameda de Osuna).

Each neighbourhood's geometric centroid is computed and the straight-line (Euclidean) distance from that centroid to the centroid of the Sol neighbourhood (`016`) is measured, used as the "center" since it is kilometer 0 of the Spanish radial highways. Computing the distance from the neighbourhood's centroid was chosen for simplicity and because of its presence in the literature; however, the geometric centroid is not necessarily where people live. The coordinate system is the same in both (ETRS_1989_UTM_Zone_30N).

### Dataset cleaning

#### Computing the indicator in 2017 and 2021

The type of planar (x,y) coordinate system used by both 2017 and 2021 in their data is made explicit. The 2017 .prj file declares a compound CRS (with a vertical z component) that the polygon geometry doesn't actually use. It is homogenized with 2021 in order to compare centroids across the two dates without sf trying to reproject or failing due to a metadata mismatch.

```{r limpieza 16, warning=FALSE}
# 2017

barrios_shp_2017 <- st_read("../data/raw/geometria_barrios_extracted/Barrios2017/Barrios.shp",
                             options = "ENCODING=CP1252", quiet = TRUE)
st_crs(barrios_shp_2017) <- 25830 # ETRS_1989_UTM_Zone_30N

# computes each neighbourhood's centroid
barrios_centroides_2017 <- barrios_shp_2017 %>%
  mutate(centroide = st_centroid(geometry))

# determines the neighbourhood centroid used as the center reference in 2017: Sol
centro_ref_2017 <- barrios_centroides_2017 %>% filter(CODBAR == "016") %>% pull(centroide)

# computes the distance between the neighbourhoods' centroids and Sol's with st_distance
distancia_centroide_barrio_2017 <- barrios_centroides_2017 %>%
  st_drop_geometry() %>%
  mutate(distancia_centro_lineal_m = as.numeric(st_distance(centroide, centro_ref_2017))) %>%
  select(codigo_barrio = CODBAR, distancia_centro_lineal_m) #CODBAR


# 2021
barrios_shp_2021 <- st_read("../data/raw/geometria_barrios_extracted/Barrios2021/Barrios_20210712.shp", quiet = TRUE)
st_crs(barrios_shp_2021) <- 25830 # ETRS_1989_UTM_Zone_30N

# computes each neighbourhood's centroid
barrios_centroides_2021 <- barrios_shp_2021 %>%
  mutate(centroide = st_centroid(geometry))

# determines the neighbourhood centroid used as the center reference in 2021: Sol
centro_ref_2021 <- barrios_centroides_2021 %>% filter(COD_BAR == "016") %>% pull(centroide)

# computes the distance between the neighbourhoods' centroids and Sol's with st_distance
distancia_centroide_barrio_2021 <- barrios_centroides_2021 %>%
  st_drop_geometry() %>%
  mutate(distancia_centro_lineal_m = as.numeric(st_distance(centroide, centro_ref_2021))) %>%
  select(codigo_barrio = COD_BAR, distancia_centro_lineal_m) #COD_BAR
```

#### Broadcast to the study years (two segments)

The broadcast is done in parts: years 2017-2020 use the 2017 vintage, years 2021-2023 use the 2021 one. Since the 2017 shapefile has 131 neighbourhoods it can be used for the 2017-2020 period.

```{r diagnostico 16}
# diagnosis of the need for two segments when doing the broadcast -> some neighbourhoods change boundary (and therefore centroid) between 2017 and 2021

# investigates which neighbourhoods changed, comparing each neighbourhood's centroid in 2017 and 2021
comparacion_centroides <- barrios_centroides_2017 %>%
  st_drop_geometry() %>%
  transmute(codigo_barrio = CODBAR, centroide_2017 = centroide) %>%
  left_join(barrios_centroides_2021 %>% # joins 2017 and 2021
      st_drop_geometry() %>%
      transmute(codigo_barrio = COD_BAR, centroide_2021 = centroide), by = "codigo_barrio") %>%
  mutate(desplazamiento_m = as.numeric(st_distance(centroide_2017, centroide_2021, by_element = TRUE))) # computes the distance between both years' centroids / by_element = TRUE makes it compare only the corresponding pairs row by row

# checks how many neighbourhoods have had their centroid shift by more than 10 meters, and it is only 4 neighbourhoods
# 10 meters is chosen to avoid capturing small digitization differences between versions
barrios_cambio_limite <- comparacion_centroides %>%
  filter(desplazamiento_m > 10) %>%
  pull(codigo_barrio)
barrios_cambio_limite

stopifnot(setequal(barrios_cambio_limite, c("214", "166", "161", "211")))
```

```{r broadcast 16}
# broadcast
# there are two map_dfr() calls applying the function, one for each segment, stacked together with bind_rows()
distancia_centro_lineal_barrio_anio <- bind_rows(
  map_dfr(2017:2020, function(anio_actual) {
    distancia_centroide_barrio_2017 %>% mutate(anio = anio_actual, .before = 1)
  }),
  map_dfr(2021:2023, function(anio_actual) {
    distancia_centroide_barrio_2021 %>% mutate(anio = anio_actual, .before = 1)
  }))
```

### Checks (02).

```{r compro 16}
# checks that both shapefiles have 131 neighbourhoods and no duplicate codes
stopifnot(n_distinct(barrios_shp_2017$CODBAR) == 131)
stopifnot(n_distinct(barrios_shp_2021$COD_BAR) == 131)
stopifnot(sum(duplicated(barrios_shp_2017$CODBAR)) == 0)
stopifnot(sum(duplicated(barrios_shp_2021$COD_BAR)) == 0)

# checks that the universe of neighbourhood codes matches barrios_madrid in both years
stopifnot(setequal(barrios_shp_2017$CODBAR, barrios_madrid$codigo_barrio))
stopifnot(setequal(barrios_shp_2021$COD_BAR, barrios_madrid$codigo_barrio))

# confirms that both coordinate systems are projected (meters), not lat/long
stopifnot(st_is_longlat(barrios_shp_2017) == FALSE)
stopifnot(st_is_longlat(barrios_shp_2021) == FALSE)

# checks that the centroid falls within its own polygon in all 131 cases, in both years
stopifnot(all(st_within(barrios_centroides_2017$centroide, barrios_centroides_2017$geometry, sparse = FALSE) %>% diag()))
stopifnot(all(st_within(barrios_centroides_2021$centroide, barrios_centroides_2021$geometry, sparse = FALSE) %>% diag()))

# checks there are no NAs in both years
stopifnot(sum(is.na(distancia_centroide_barrio_2017$distancia_centro_lineal_m)) == 0)
stopifnot(sum(is.na(distancia_centroide_barrio_2021$distancia_centro_lineal_m)) == 0)

# checks the number of rows in the final table
stopifnot(nrow(distancia_centro_lineal_barrio_anio) == 131 * 7)

# checks that Sol gives distance 0 in both years
stopifnot(distancia_centroide_barrio_2017$distancia_centro_lineal_m[distancia_centroide_barrio_2017$codigo_barrio == "016"] == 0)
stopifnot(distancia_centroide_barrio_2021$distancia_centro_lineal_m[distancia_centroide_barrio_2021$codigo_barrio == "016"] == 0)

# checks that the broadcast is correct within each year segment (2017-2020 and 2021-2023)
chk_broadcast_centro_2017 <- distancia_centro_lineal_barrio_anio %>%
  filter(anio <= 2020) %>%
  group_by(codigo_barrio) %>%
  summarise(n_valores_distintos = n_distinct(distancia_centro_lineal_m), .groups = "drop")
stopifnot(all(chk_broadcast_centro_2017$n_valores_distintos == 1))

chk_broadcast_centro_2021 <- distancia_centro_lineal_barrio_anio %>%
  filter(anio >= 2021) %>%
  group_by(codigo_barrio) %>%
  summarise(n_valores_distintos = n_distinct(distancia_centro_lineal_m), .groups = "drop")
stopifnot(all(chk_broadcast_centro_2021$n_valores_distintos == 1))
```

### Saving

```{r save 16}
write_csv(distancia_centro_lineal_barrio_anio,"../data/processed/distancia_centro_lineal_barrio_anio.csv")
```

### View of the final table

```{r view 16}
#| eval: false
distancia_centro_lineal_barrio_anio <- read_csv("../data/processed/distancia_centro_lineal_barrio_anio.csv", show_col_types = FALSE)
View(distancia_centro_lineal_barrio_anio)
```

## Source 17: unemployment rate (`tasa_paro.xlsx`)

**Dataset comments**:

This dataset is at neighbourhood level and in wide format. From this source only the absolute registered-unemployment-rate indicator, for December of each year, is needed.

### Dataset cleaning

```{r limpieza 17, message=FALSE}
raw_paro <- read_excel("../data/raw/tasa_paro.xlsx", sheet = 1, col_names = FALSE) # not assuming a header

# row 5 = fila_ciudad_paro carries the district name and the following row carries the district-total or specific-neighbourhood label
stopifnot(as.character(raw_paro[[5, 5]]) == "Ciudad de Madrid")
fila_ciudad_paro <- 5
col_ciudad_paro <- 5
etiquetas_geo_paro <- str_squish(as.character(unlist(raw_paro[fila_ciudad_paro + 1, col_ciudad_paro:ncol(raw_paro)])))

# keeps only the columns that have the neighbourhood pattern (3 digits + period) - discards "Ciudad de Madrid" and the 21 district-total columns
es_barrio_col_paro <- str_detect(etiquetas_geo_paro, "^\\d{3}\\.\\s") # 3-digit pattern
cols_barrio_paro <- (col_ciudad_paro:ncol(raw_paro))[es_barrio_col_paro]
barrio_txt_cols_paro <- etiquetas_geo_paro[es_barrio_col_paro]
stopifnot(length(cols_barrio_paro) == 131) # must be the 131 neighbourhoods



# cleaning function
parse_tasa_paro <- function(x) {
  x <- if_else(x == "..", NA_character_, x) # statistical-confidentiality sentinel becomes NA
  as.numeric(str_replace(x, ",", ".")) # rates use the European decimal comma (, is changed to .)
}

# locates the data rows: the first column carries the year as text (2017-2023)
filas_datos_paro <- which(str_detect(as.character(raw_paro[[1]]), "^\\d{4}$")) # the rows are the ones matching that year pattern
stopifnot(length(filas_datos_paro) == 7) # 2017-2023

tasa_paro_largo <- map_dfr(filas_datos_paro, function(f) { # builds the final table
  tibble(
    anio = as.integer(raw_paro[[f, 1]]),
    barrio_txt = barrio_txt_cols_paro,
    tasa_paro = parse_tasa_paro(as.character(unlist(raw_paro[f, cols_barrio_paro]))))
})


# join
tasa_paro_barrio_anio <- tasa_paro_largo %>%
  mutate(codigo_barrio = str_extract(barrio_txt, "^\\d{3}")) %>% # extracts the neighbourhood code
  left_join(barrios_madrid %>% select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio), by = "codigo_barrio") %>% # joins this table with barrios_madrid, bringing in the rest of the neighbourhood information
  select(anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio, tasa_paro) %>% arrange(anio, distrito_num, barrio_num)
```

### Checks (02).

```{r compro 17}
# checks that no codigo_barrio was left without finding its neighbourhood when joining with barrios_madrid
stopifnot(sum(is.na(tasa_paro_barrio_anio$barrio_nombre)) == 0)

# checks the number of neighbourhoods per year: 131 in all 7 years (this source doesn't have the 128-in-2017 exception, see the dataset comment)
stopifnot(all(count(tasa_paro_barrio_anio, anio)$n == 131))

# checks there are no duplicate neighbourhoods
stopifnot(sum(duplicated(tasa_paro_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# checks that 2017-2023 have no unexpected NAs
stopifnot(sum(is.na(tasa_paro_barrio_anio$tasa_paro)) == 0)
```

### Saving

```{r save 17}
write_csv(tasa_paro_barrio_anio, "../data/processed/tasa_paro_barrio_anio.csv")
```

### View of the final table

```{r view 17}
#| eval: false
tasa_paro_barrio_anio <- read_csv("../data/processed/tasa_paro_barrio_anio.csv", show_col_types = FALSE)
View(tasa_paro_barrio_anio)
```

## Final union: complete neighbourhood-year panel

**Comments**:

The 16 thematic sources (Sources 2-17) are combined into a single neighbourhood-year panel. The final panel uses 131 neighbourhoods in 2017 since several sources do have real data for Ensanche de Vallecas/Valderrivas/El Cañaveral in 2017. As for column collisions between sources, the redundant copy is dropped before joining. This dataset includes variables that will ultimately not be used but were part of the construction/decision process.

```{r grid maestro}
# Builds the panel's skeleton/grid: 131 neighbourhoods x 7 years = 917 rows
panel_final_barrio_anio <- map_dfr(2017:2023, function(anio_actual) {
  barrios_madrid %>%
    select(distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio, barrio_oficial) %>%
    mutate(anio = anio_actual, .before = 1)
})
```

```{r union fuentes}
# a list of each source's objects
# before putting them in the list, the duplicate column is removed where applicable
tablas_a_unir <- list(
  composicion_hogares_barrio_anio,
  alquiler_barrio_anio,
  precio_vivienda_declarado_barrio_anio,
  rentas_barrio_anio,
  tramo_edad_barrio_anio,
  transacciones_vivienda_barrio_anio,
  panel_indicadores_barrio_anio,
  censo_locales_barrio_anio,
  churn_total_barrio_anio %>%
    select(-poblacion_total_barrio), # duplicate of tramo_edad
  inversiones_ayto_distritos_barrio_anio,
  inversiones_tipo_distrito_barrio_anio %>%
    select(-presupuesto_gasto_distrito, -gasto_real_distrito), # duplicate of inversiones_ayto
  nacionalidad_pais_distrito_barrio_anio,
  zonas_verdes_distrito_barrio_anio,
  distancia_metro_barrio_anio %>% select(-barrio_oficial), # already in the grid
  distancia_centro_lineal_barrio_anio,
  tasa_paro_barrio_anio)

# join loop
for (tabla in tablas_a_unir) {

  # before each join, the neighbourhood metadata columns already present in the grid are dropped
  tabla_limpia <- tabla %>%
    select(-any_of(c("distrito_num", "distrito_nombre", "barrio_num", "barrio_nombre"))) # any_of allows removing names that don't exist in every table without throwing an error
  panel_final_barrio_anio <- panel_final_barrio_anio %>%
    left_join(tabla_limpia, by = c("anio", "codigo_barrio"))

  # check
  cols_nuevas_sospechosas <- names(panel_final_barrio_anio)[str_detect(names(panel_final_barrio_anio), "\\.(x|y)$")] # detects whether there are columns with .x .y suffixes (duplicates)
  if (length(cols_nuevas_sospechosas) > 0) { # if those columns exist, it reports where and stops execution
    stop(sprintf("Unexpected column-name collision while joining this source: %s",
                 paste(cols_nuevas_sospechosas, collapse = ", ")))
  }}
```

### Checks (02).

```{r compro union}
# checks the panel's row count: 131 neighbourhoods x 7 years
stopifnot(nrow(panel_final_barrio_anio) == 131 * 7)

# the join keys (anio, codigo_barrio) must have no NA nor duplicates
stopifnot(sum(is.na(panel_final_barrio_anio$anio)) == 0)
stopifnot(sum(is.na(panel_final_barrio_anio$codigo_barrio)) == 0)
stopifnot(sum(duplicated(panel_final_barrio_anio %>% select(anio, codigo_barrio))) == 0)

# column count per source, to visually check that no entire source is missing
cat("Total number of columns in the final panel:", ncol(panel_final_barrio_anio), "\n")

# each table in the list is inspected before starting to join, in case any column carries the .x .y suffixes (which left_join uses when a column name is shared)
for (tabla in tablas_a_unir) {
  cols_sospechosas <- names(tabla)[str_detect(names(tabla), "\\.(x|y)$")]
  if (length(cols_sospechosas) > 0) {
    stop(sprintf("Found .x/.y columns (%s) -- rebuild this source (Restart R + Run All) before joining", paste(cols_sospechosas, collapse = ", ")))
  }}
```

### Saving

```{r save union}
write_csv(panel_final_barrio_anio, "../data/processed/panel_final_barrio_anio.csv")
```

### View of the final table

```{r view union}
#| eval: false
panel_final_barrio_anio <- read_csv("../data/processed/panel_final_barrio_anio.csv", show_col_types = FALSE)
View(panel_final_barrio_anio)
```

## Final table with the selected variables

**Comments**:

This dataset is the closed selection of the index and predictor variables. `pct_hog_unipersonal_bajo65`, `pct_hog_monoparental` and `pct_hog_dos_adultos_sin_menores` are computed here from columns already present in `panel_final_barrio_anio`. `pct_extranjeros_ue_ocde`

```{r final}
dataset_modelado <- panel_final_barrio_anio %>%
  mutate(
    pct_hog_unipersonal_bajo65 = (hog_mujer_sola_16_64 + hog_hombre_solo_16_64) / total_hogares,
    pct_hog_monoparental = (hog_mujer_adulta_con_menores + hog_hombre_adulto_con_menores) / total_hogares,
    pct_hog_dos_adultos_sin_menores = hog_dos_adultos_sin_menores / total_hogares) %>%
  
  select(
    # identifiers
    anio, distrito_num, distrito_nombre, barrio_num, barrio_nombre, codigo_barrio,

    # index (6 dimensions that also enter as a t0-level predictor)
    pct_pob_25_39, nivel_estudios_superiores_pct, renta_neta_media_persona,
    pct_hog_unipersonal_bajo65, precio_vivienda_declarado_barrio,
    pct_extranjeros_ue_ocde,

    # predictors
    pct_hog_monoparental, pct_hog_dos_adultos_sin_menores,
    renta_colectiva_barrio,
    poblacion_total_barrio, tasa_transacciones_vivienda, densidad, tamano_hogar,
    envejecimiento, sobreenvejecimiento, pct_uso_vivienda,
    n_locales_activos_total, pct_comercio_menor, pct_turismo_alojamiento, pct_bares_restaurantes,
    pct_inmobiliarias, pct_arte_recreo_ocio, tasa_entradas_totales, tasa_salidas_totales,
    intensidad_inversion_real_distrito, share_gasto_real_urbanismo, share_gasto_real_mantenimiento_urbano,
    share_gasto_real_medio_ambiente, share_gasto_real_cultura, pct_poblacion_idh_alto_distrito,
    zonas_verdes_significativas_m2_per_capita_distrito, paradas_red_centro, n_estaciones_metro,
    distancia_centro_lineal_m, tasa_paro)
```

### Saving

```{r save final}
write_csv(dataset_modelado, "../data/processed/dataset_modelado.csv")
```

### View of the final table

```{r view final}
dataset_modelado <- read_csv("../data/processed/dataset_modelado.csv", show_col_types = FALSE)
View(dataset_modelado)
```


## Selected variables summary


| Variable | Role | Source | Description |
|---|---|---|---|
| `pct_pob_25_39` | indicator and predictor | 6 | Share of the neighbourhood's population aged 25-39. |
| `nivel_estudios_superiores_pct` | indicator and predictor | 8 | Share of the population with completed higher education. |
| `renta_neta_media_persona` | indicator and predictor | 5 | Average net income per person in the neighbourhood. |
| `pct_hog_unipersonal_bajo65` | indicator and predictor | 2 | Share of single-person households with the resident aged under 65. |
| `precio_vivienda_declarado_barrio` | indicator and predictor | 4 | Average declared housing sale price registered by the Colegio de Registradores for the neighbourhood. |
| `pct_extranjeros_ue_ocde` | indicator and predictor | 8 | Share of the population holding EU/OECD foreign nationality. |
| `pct_hog_monoparental` | predictor | 2 | Share of households that are single-parent households (one adult with one or more children). |
| `pct_hog_dos_adultos_sin_menores` | predictor | 2 | Share of households with two adults aged 16-64 with no children. |
| `renta_colectiva_barrio` | predictor | 3 | Median rental price of multi-family housing in the neighbourhood, built as a dwelling-weighted mean of its census tracts' median rents. |
| `poblacion_total_barrio` | predictor | 6 | Total resident population of the neighbourhood. |
| `tasa_transacciones_vivienda` | predictor | 7 | Housing sale transactions per inhabitant in the neighbourhood. |
| `tasa_paro` | predictor | 17 | Registered unemployment rate in the neighbourhood (measured in December of each year). |
| `densidad` | predictor | 8 | Population density: inhabitants per hectare of the neighbourhood's surface area. |
| `tamano_hogar` | predictor | 8 | Average household size (mean number of people per household). |
| `envejecimiento` | predictor | 8 | Population aging ratio (share of elderly population >65)|
| `sobreenvejecimiento` | predictor | 8 | Over-aging ratio (share of the population >80 within the elderly population) |
| `pct_uso_vivienda` | predictor | 9 | Share of all registered business/premises whose declared status is residential use |
| `n_locales_activos_total` | predictor | 9 | Total number of open (active) business premises in the neighbourhood. |
| `pct_comercio_menor` | predictor | 9 | Share of active business premises devoted to retail trade ("Comercio al por menor"). |
| `pct_turismo_alojamiento` | predictor | 9 | Share of active business premises devoted to tourist accommodation (hotels and similar and tourist housing). |
| `pct_bares_restaurantes` | predictor | 9 | Share of active business premises devoted to bars and restaurants. |
| `pct_inmobiliarias` | predictor | 9 | Share of active business premises devoted to real-estate agencies. |
| `pct_arte_recreo_ocio` | predictor | 9 | Share of active business premises devoted to arts, recreation and leisure activities. |
| `tasa_entradas_totales` | predictor | 10 | Rate of new municipal entries (internal + external "altas") per inhabitant, capturing in-migration into the neighbourhood. |
| `tasa_salidas_totales` | predictor | 10 | Rate of municipal de-registrations (internal + external "bajas") per inhabitant, capturing out-migration from the neighbourhood. |
| `intensidad_inversion_real_distrito` | predictor | 11 | District-level actual municipal spending execution per inhabitant|
| `share_gasto_real_urbanismo` | predictor | 12 | Share of the district's actual municipal spending allocated to urbanism |
| `share_gasto_real_mantenimiento_urbano` | predictor | 12 | Share of the district's actual municipal spending allocated to urban maintenance |
| `share_gasto_real_medio_ambiente` | predictor | 12 | Share of the district's actual municipal spending allocated to the environment |
| `share_gasto_real_cultura` | predictor | 12 | Share of the district's actual municipal spending allocated to culture |
| `pct_poblacion_idh_alto_distrito` | predictor | 13 | Share of the district's population born abroad in a country with high or very high HDI (>0.7) |
| `zonas_verdes_significativas_m2_per_capita_distrito` | predictor | 14 | District-level surface area of significant parks and green spaces per inhabitant (m2 per capita). |
| `paradas_red_centro` | predictor | 15 | Minimum number of metro-network segments from the neighbourhood's own station(s) to Sol along the shortest path, used as a network-distance-to-center measure |
| `n_estaciones_metro` | predictor | 15 | Number of metro stations located within the neighbourhood's boundaries. |
| `distancia_centro_lineal_m` | predictor | 16 | Straight-line (Euclidean) distance, in meters, from the neighbourhood's geometric centroid to the centroid of Sol. |









